# 2024.10.01.
# Anna Lovrics

---
title: "Ras interactome analysis"
date: "`r Sys.Date()`"
author: "Anna Lovrics"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
---

## Aim of this analysis is to find best genes that seperate normal from
tumor or even normal from stage I samples. 

```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```


```{r libraries}
library(readxl) # another package to read xls files
library(dplyr)
library(tidyr) # to transform data for ggplot2 plots
library(data.table) # while I use tibble mostly, fread is best input option
library(tibble)
library(ggplot2) # for visualization
library(EnvStats) # for statistic tests
library(ggcorrplot) # for nice correlation plots
library(cowplot) # to plot q-q plots in a grid
library(gtools) # specifically to generate stars from significance values
library(caret) # for Classification And REgression Training
library(pROC) # to easily plot ROC curves
library(limma) # for the strsplit2 function
library(gridExtra) # to plot a list of figures in a grid
library(ggpubr) # for the same purpose to arrange figures in a grid
```

# Folder structure
```{r folders}
phome <- "/media/anna/bigdisk/Anna/Enzimhome_local/Projects/StatisticsCore/Ras"
datafolder <- file.path(phome, "Data")
figfolder <- file.path(phome, "Figures")
if (!file.exists(figfolder)) dir.create(figfolder)
calcfolder <- file.path(phome, "Data_calculated")
if (!file.exists(calcfolder)) dir.create(calcfolder)
resfolder <- file.path(phome, "Results")
# source content of utility file
rscriptfolder <- file.path(phome, "Rscript")
util_file <- file.path(rscriptfolder, "anal_utils.R")
source(util_file)
# also attach ppi utils
ppi_utils_file <- file.path("/media/anna/bigdisk/Anna/Enzimhome_local",
      "Projects/PPi/Rscript/ppi0_utils.R")
source(ppi_utils_file)
```

# read in result from all cancer types
```{r input data}
ras_data_list <- list()
for (rtype in c("BRCA", "COAD", "LUAD", "LUSC")){
  # read data
  ras_list <- readRDS(file=file.path(calcfolder, rtype,
                                     paste("TCGA_", rtype, ".rds", sep="")))
  ras_data_list[[rtype]] <- ras_list[["data"]]
  indep_vars <- ras_list[["indep_vars"]]
}

```


# create clusters based on correlation between variables
```{r correlation}
cl_list <- list()
for (rtype in names(ras_data_list)){
  corr <- cor(ras_data_list[[rtype]][, indep_vars], 
                  use="pairwise.complete.obs",
                  method = "pearson")
  # create clusters
  hc <- hclust(dist(corr), method="ward.D2")
  cl_list[[rtype]] <- cutree(hc, k=5)
  # sort by alphabetical order
  cl_list[[rtype]] <- cl_list[[rtype]][order(names(cl_list[[rtype]]))]
}
```

```{r calculate co-association matrix}
# Initialize an empty co-association matrix
co_assoc_matrix <- matrix(0, nrow = length(indep_vars), 
                          ncol = length(indep_vars))

# Fill the co-association matrix based on clustering results
for (clustering in cl_list) {
  # Create a pairwise indicator matrix for the current clustering
  indicator_matrix <- outer(clustering, clustering, FUN = "==")
  
  # Add this indicator matrix to the co-association matrix
  co_assoc_matrix <- co_assoc_matrix + indicator_matrix
}

# Convert counts to proportions (similarity matrix)
co_assoc_matrix <- co_assoc_matrix / length(cl_list)

# apply clustering on the co-association matrix
hc_all <- hclust(as.dist(1 - co_assoc_matrix), method="ward.D2")
final_clusters <- cutree(hc_all, k=5)
```

```{r plot correlations in the order of the final clusters}
plist <- list()
cluster_corr_list <- list()
for (rtype in names(ras_data_list)){
  corr <- cor(ras_data_list[[rtype]][, indep_vars], 
                  use="pairwise.complete.obs",
                  method = "pearson")
  norder <- names(sort(final_clusters))
  corder <- match(norder, colnames(corr))
  corr <- round(corr[corder, corder], digits=1)
  # add cluster number to labels
  newlabels <- paste(sort(final_clusters), colnames(corr), sep="_")
  rownames(corr) <- newlabels
  colnames(corr) <- newlabels
  cluster_corr_list[[rtype]] <- corr
  plist[[rtype]] <- ggcorrplot(corr, lab=T, hc.order=F)
  figfile <- file.path(figfolder, rtype,
                          paste("final_clusters_", rtype, ".pdf", sep=""))
  ggsave(figfile, plot=plist[[rtype]], width=20, height =20)
}

# also plot all together
pfinal <- cowplot::plot_grid(plotlist=plist, ncol=2)
figfile <- file.path(figfolder, "final_clusters_all.pdf")
ggsave(figfile, plot=pfinal, width=20, height=20)

```

```{r color labels by best model variables}

# read in best models
best_models <- list()
plist <- list()
for (rtype in names(ras_data_list)){
  infile <- file.path(resfolder, 
                paste("best_models_", rtype, ".txt", sep=""))
  best_models[[rtype]] <- read.delim(infile, header=F)[[1]]
  corr <- cluster_corr_list[[rtype]]
  # 
  # # create a long tibble
  # corr_data <- reshape2::melt(corr)
  # colnames(corr_data) <- c("Var1", "Var2", "Correlation")

  # # color with red if in best models
  # category <- rep("NO", times=length(indep_vars))
  # names(category) <- indep_vars
  # category[names(category) %in% best_models[[rtype]]] <- "YES"
  # 
  # # Define a color scheme for each category
  # category_colors <- c("NO" = "black", "YES" = "red")
  
  # color with red if in best models
  category <- rep("NO", times=length(indep_vars))
  names(category) <- indep_vars
  category[names(category) %in% best_models[[rtype]]] <- "YES"
  # now replace category names
  orig_names <- limma::strsplit2(colnames(corr), split="_")[,2]
  dict_names <- cbind("original"=orig_names, "new"=colnames(corr))
  names(category) <- dict_names[, "new"][match(names(category), 
                                          dict_names[, "original"])]
  # order as in corr
  category <- category[match(colnames(corr), names(category))]
  
  # Define a color scheme for each category
  category_colors <- c("NO" = "black", "YES" = "red")
  
  # Create a data frame to map each variable to its color
  # label_data <- data.frame(
  #   variable = colnames(corr),
  #   color = category_colors[category[colnames(corr)]]
  # )
  # 
  # # Create the heatmap without axis text
  # #p <- ggplot(corr_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  # p <- ggplot(corr_data, aes(x = Var1, y = Var2)) +
  #   geom_tile(color = "white")+    # Heatmap tiles with white borders
  #   # scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
  #   #                      midpoint = 0, limit = c(-1, 1), 
  #   #                      name = "Correlation") +
  #   theme_minimal() +
  #   theme(
  #     axis.text.x = element_blank(),
  #     axis.text.y = element_blank(),
  #     axis.ticks = element_blank()
  #   )
  # 
  # # add colored labels
  # p <- p + 
  #   geom_text(data = label_data, 
  #             aes(x = variable, y = -0.5, label = variable, color = color), 
  #             angle = 45, hjust = 1, size = 5)+
  #    geom_text(data = label_data, 
  #           aes(x = -0.5, y = variable, label = variable, color = color), 
  #           hjust = 1, size = 5) +
  #     scale_color_identity() +
  #     coord_fixed(ratio = 1)
  
  plist[[rtype]] <- ggcorrplot(corr, lab = TRUE) + 
  theme(axis.text.x = element_blank(), axis.text.y = element_blank()) +
  scale_x_discrete(
    labels = names(category),
    expand = expansion(add = 0.5)
  ) +
  theme(
    axis.text.x = element_text(
      color = sapply(names(category), function(name) category_colors[category[name]]),
      angle = 45, hjust = 1, size = 10
    )
  ) +
  theme(
    axis.text.y = element_text(
      color = sapply(names(category), function(name) category_colors[category[name]]),
      size = 10
    )
  )

  
}

# also plot all together
pfinal <- cowplot::plot_grid(plotlist=plist, ncol=2, labels=names(plist))
figfile <- file.path(figfolder, "final_clusters_all_colored.pdf")
ggsave(figfile, plot=pfinal, width=30, height=30)

```


