# 2024.10.01.
# Anna Lovrics

# TO DO:
# 1) 3-6 genes balanced accuracy  (melyik milyen irányba megy?, pl)
# 2) other cancer: best 3-gene confusion matrix & balanced accuracy

---
title: "Ras interactome analysis"
date: "`r Sys.Date()`"
author: "Anna Lovrics"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```


```{r libraries}
library(readxl) # another package to read xls files
library(dplyr)
library(tidyr) # to transform data for ggplot2 plots
library(data.table) # while I use tibble mostly, fread is best input option
library(tibble)
library(ggplot2) # for visualization
library(EnvStats) # for statistic tests
library(ggcorrplot) # for nice correlation plots
library(cowplot) # to plot q-q plots in a grid
library(gtools) # specifically to generate stars from significance values
library(caret) # for Classification And REgression Training
library(pROC) # to easily plot ROC curves
library(limma) # for the strsplit2 function
library(gridExtra) # to plot a list of figures in a grid
library(ggpubr) # for the same purpose to arrange figures in a grid and also for significance stars
library(arsenal) # Functions for Large-Scale Statistical Summaries
```

# Folder structure
```{r folders}
phome <- "/media/anna/bigdisk/Anna/Enzimhome_local/Projects/StatisticsCore/Ras"
datafolder <- file.path(phome, "Data")
figfolder <- file.path(phome, "Figures")
resfolder <- file.path(phome, "Results")
if (!file.exists(figfolder)) dir.create(figfolder)
calcfolder <- file.path(phome, "Data_calculated")
if (!file.exists(calcfolder)) dir.create(calcfolder)
# source content of utility file
rscriptfolder <- file.path(phome, "Rscript")
util_file <- file.path(rscriptfolder, "anal_utils.R")
source(util_file)
```

# define the cancer type
```{r cancer type}
#rtype <- "COAD"
rtype <- "COAD"
figfolder <- file.path(figfolder, rtype)
if (!file.exists(figfolder)) dir.create(figfolder)
calcfolder <- file.path(calcfolder, rtype)
if (!file.exists(calcfolder)) dir.create(calcfolder)
resfolder <- file.path(resfolder, rtype)
if (!file.exists(resfolder)) dir.create(resfolder)
```

# read in data
```{r read data, results="asis"}
ras_data <- fread(file.path(datafolder, 
                  paste("TCGA_", rtype, "_plus_stage_20241028.tsv", sep="")),
                     sep="\t")
ras_data <- as_tibble(ras_data)
# remove FFPE Scrolls sample
tmpi <- which(ras_data$sample_type=="FFPE Scrolls")
if (length(tmpi) > 0) ras_data <- ras_data[-tmpi,]
# exclude some of the genes
exclude <- c("RAPGEF3", "RAPGEF5", "RASIP1", "MYO9B", "PIK3C2A", "RGL3", "RGL4",
             "TIAM1", "ARHGAP20","RASSF2", "RASSF3", "RASSF4", "RASSF6")
keep_cols <- setdiff(colnames(ras_data), exclude)
ras_data <- ras_data[, keep_cols]
# View(ras_data)
# first two columns are the same -> remove
# all.equal(ras_data$sample, ras_data$samples)
ras_data$samples <- NULL
indep_vars <- setdiff(colnames(ras_data), 
                      c( "sample","pathologic_stage","sample_type"))
# keep original data, replace columns with what we need in the computations
ras_data_orig <- ras_data
# add a column for tumor or normal
ras_data$sample_type2 <- ras_data$sample_type
ras_data$sample_type2[ras_data$sample_type2 %in% 
    c("Metastatic", "Primary Tumor", "Recurrent Tumor")] <- "Tumor"
ras_data$sample_type2[ras_data$sample_type2=="Solid Tissue Normal"] <- "Normal"
# add a column for stages I-IV or X
ras_data$pathologic_stage2 <- ras_data$pathologic_stage
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage I","Stage IA", "Stage IB", "Stage IC")] <- "Stage_I"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage II","Stage IIA", "Stage IIB", "Stage IIC")] <- "Stage_II"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage III","Stage IIIA", "Stage IIIB", "Stage IIIC")] <- "Stage_III"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage IV", "Stage IVA", "Stage IVB", "Tumor_Stage IVB")] <- "Stage_IV"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("[Discrepancy]","Stage X")] <- "Stage_X"

# sample type plusz stage together
ras_data$category <- paste(ras_data$sample_type2, ras_data$pathologic_stage2, sep="_")
ras_data$category2 <- ras_data$category
ras_data$category2[ras_data$category2 %in% c("Normal_Stage_X", "Normal_Stage_I",
     "Normal_Stage_II", "Normal_Stage_III", "Normal_Stage_IV")] <- "Normal"

# check for constant columns and remove if any
cvar <- apply(ras_data[, indep_vars], 2, var, na.rm=T)
ci <- which(cvar==0)
if (length(ci)>0) ras_data <- ras_data[-ci,]

# add some more response variables
ras_data$sample_type_factor <- factor(ras_data$sample_type2,
                                 levels=c("Normal", "Tumor"))
ras_data$sample_type_bool <- ifelse(ras_data$sample_type2=="Tumor", F, T)
ras_data$category_factor <- factor(ras_data$category2)
# save data
saveRDS(list("data"=ras_data, "indep_vars"=indep_vars),
        file=file.path(calcfolder, paste("TCGA_", rtype, ".rds", sep="")))

# # print out tables for check
# print(table(ras_data$sample_type_factor))
# print(table(ras_data$category_factor))
# xtable::xtable(table(ras_data$category_factor))

tab1 <- tableby( ~ category_factor, data=ras_data)
# summary(tab1)
# Create HTML output
html_output <- summary(tab1, text = FALSE)
outfile <- file.path(resfolder, "stages_summary.docx")
write2word(summary(tab1), outfile)
outfile <- gsub(".docx", ".html", outfile)
write2html(summary(tab1), outfile, title="summary")
#browseURL(outfile)
```
# boxplots and histograms
```{r create boxplots and histograms}
# create long data for ggplot visualizations
ras_long <- ras_data %>% 
       pivot_longer(all_of(indep_vars), 
                    names_to = "gene", values_to = "expression")
# to keep variables order:
ras_long$gene <- factor(ras_long$gene, levels=indep_vars)

# plot boxplots for all independent variables
# also add result of t-test
pvalue_list <- list()
signif_list <- list()
for (ivar in indep_vars){
  ttest_res <- (
    t.test(ras_data[[ivar]][ras_data$sample_type2=="Normal"],
           ras_data[[ivar]][ras_data$sample_type2=="Tumor"]))
  pvalue_list[[ivar]] <- ttest_res$p.value
  signif_list[[ivar]] <- stars.pval(ttest_res$p.value)
}
best_indep <- names(which(pvalue_list==min(unlist(pvalue_list))))
pvalue_tbl <- as_tibble(pvalue_list)
pvalue_long <- pivot_longer(pvalue_tbl, cols = everything(),
                            names_to="gene", values_to="pvalue")
pvalue_long <- pvalue_long[order(pvalue_long$pvalue),] # sort
# # create a barplot
# scaling_tbl <- scaling_tbl[order(scaling_tbl$LD2, decreasing = F),]
# scaling_tbl$gene <- factor(scaling_tbl$gene, levels=scaling_tbl$gene)
# p2 <- ggplot(scaling_tbl) +
#   geom_col(aes(x=gene, y=LD2), fill = "red", width = 0.6)+
#   coord_flip()
# p <- cowplot::plot_grid(p1, p2, nrow=1)
# outfig <- file.path(figfolder, "lda_scaling.pdf")
# ggsave(outfig, plot=p, width=10, height=6)

signif_tbl <- as_tibble(signif_list)
signif_long <- pivot_longer(signif_tbl, cols = everything(),
                            names_to="gene", values_to="significance")
# to keep variables order:
signif_long$gene <- factor(signif_long$gene,  levels=indep_vars)
signif_long$sample_type2 <- "Tumor" # so that asterix is above tumor
signif_long$yval <- 0
for (ivar in unique(ras_long$gene)){
  maxval <- median(ras_data[[ivar]]) + 1.5*IQR(ras_data[[ivar]])
  signif_long[signif_long$gene==ivar, "yval"] <- 1.1*maxval
}

# plot finally
p <- ggplot(data=ras_long, aes(x=sample_type2, y=expression)) +
  #geom_boxplot() + 
  geom_violin(trim=F) +
  geom_jitter(width = 0.1, alpha = 0.3, size=0.1) +
  geom_boxplot(width=.1)+
  geom_text(data=signif_long, aes(label=significance, y=yval),
            fontface=2) +
  facet_wrap(~gene, scale="free")
#figfile <- file.path(figfolder, "boxplot_all.pdf")
figfile <- file.path(figfolder, "violinplot_all.pdf")
ggsave(figfile, p, width=14, height=14)

# also show the boxplot for different stages
# plot finally
p <- ggplot(data=ras_long[!ras_long$category_factor=="Tumor_Stage_X",], 
            aes(x=category2, y=expression)) +
  geom_boxplot() + 
  facet_wrap(~gene, scale="free", ncol=4)+
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
figfile <- file.path(figfolder, "boxplot_stages.pdf")
ggsave(figfile, p, width=14, height=42)

# plot histograms for all independent variables
ras_long_tmp <- ras_long
ras_long_tmp$sample_type2 <- factor(ras_long_tmp$sample_type2,
                                       levels=c("Tumor", "Normal"))
p <- ggplot(data=ras_long_tmp, 
            mapping=aes(x=expression, fill=sample_type2))+
  geom_histogram()+
  facet_wrap(~gene, scales="free")
suppressWarnings(plot(p))
figfile <- file.path(figfolder, "histogram_all.pdf")
suppressWarnings(ggsave(figfile, p, width=14, height=14))
```

# Correlation between variables
```{r correlation}
set.seed(1004)
cfigfolder <- file.path(figfolder, "correlation")
if (!file.exists(cfigfolder)) dir.create(cfigfolder)
# plot correlation among the variables
# do it for all samples, tumor samples and normal samples
# order always as would be for normal
corr <- round(cor(ras_data[which(ras_data$sample_type2=="Normal"), indep_vars], 
                  use="pairwise.complete.obs",
                  method = "pearson"), digits=1)
# order by the "Normal samples clustering
hc <- hclust(dist(corr), method="ward.D2")
# use the number of samples that are minimal in the stages
slist <- setdiff(unique(ras_data$category2), "Tumor_Stage_X")
num_samples_min <- min(table(ras_data[ras_data$category2 %in% slist,]$category2))
# now plot
corr_list <- list()
plot_list <- list()
mac_list <- list()
for (stype in c("Normal","All", "Tumor")){
  sel_samples <- 1:nrow(ras_data)
  if (stype=="Normal"){
    sel_samples <- which(ras_data$sample_type2=="Normal")
    sel_samples <- sample(sel_samples, size=num_samples_min)
  }
  if (stype=="Tumor"){
    sel_samples <- which(ras_data$sample_type2=="Tumor")
    sel_samples <- sample(sel_samples, size=num_samples_min)
  }
  if (stype=="All"){
    sel_samples <- caret::createDataPartition(factor(ras_data$sample_type2), 
                                      p=num_samples_min/nrow(ras_data))[[1]]
  }
  corr <- round(cor(ras_data[sel_samples, indep_vars], 
                    use="pairwise.complete.obs",
                    method = "pearson"), digits=1)
  # mean absolute correlation
  #mac_list[[stype]] <- mean(abs(corr[lower.tri(corr, diag = FALSE)]))
  mac_list[[stype]] <- length(which(abs(corr[lower.tri(corr, diag = FALSE)])>0.4))
  # #corr <- corr[hc$order, hc$order]
  # figfile <- file.path(cfigfolder, 
  #               paste("correlation_", stype, "_orig_order.pdf", sep=""))
  # p <- ggcorrplot(corr, lab=T,
  #                 hc.order=F, hc.method="ward.D2")
  # ggsave(figfile, plot=p, width=20, height =20)
  # save clustering order as well
  corr <- corr[hc$order, hc$order]
  if (stype=="Normal") stype <- "Normal_first"
  figfile <- file.path(cfigfolder, 
                       paste("correlation_", stype, "_hc_order.pdf", sep=""))
  p <- ggcorrplot(corr, lab=T,
                  hc.order=F, hc.method="ward.D2")
  ggsave(figfile, plot=p, width=20, height =20)
  # store in list
  corr_list[[stype]] <- corr
  plot_list[[stype]] <- p
}
print(mac_list)
# write out to text file as well
outfile <- file.path(resfolder, "correlation.xlsx")
openxlsx2::write_xlsx(corr_list[c("All", "Normal_first", "Tumor")], outfile, rowNames=T)


# now plot
corr_category_list <- list()
plot_category_list <- list()
slist <- setdiff(unique(ras_data$category2), "Tumor_Stage_X")
num_samples_min <- min(table(ras_data[ras_data$category2 %in% slist,]$category2))
for (stype in slist){
  sel_samples <- which(ras_data$category2==stype)
  sel_samples <- sample(sel_samples, size=num_samples_min)
  corr <- round(cor(ras_data[sel_samples, indep_vars], 
                    use="pairwise.complete.obs",
                    method = "pearson"), digits=1)
  # save clustering order as well
  corr <- corr[hc$order, hc$order]
  figfile <- file.path(cfigfolder,
                       paste("correlation_", stype, "_hc_order.pdf", sep=""))
  p <- ggcorrplot(corr, lab=T,
                  hc.order=F, hc.method="ward.D2")
  ggsave(figfile, plot=p, width=20, height =20)
  # store in list
  corr_category_list[[stype]] <- corr
  plot_category_list[[stype]] <- p
}

# create a plot with normal, tumor and stage I
plot_manuscript <- cowplot::plot_grid(plot_category_list[["Normal"]],
                                      plot_list[["Tumor"]],
                                      plot_category_list[["Tumor_Stage_I"]],
                                      ncol=1,
                                      labels = c("Normal", "Tumor", "Stage I"))
figfile <- file.path(cfigfolder, "correlation_grid.pdf")
ggsave(figfile, plot=plot_manuscript, width=20, height=60, limitsize = F)
```

<!-- # define ap clusters -->
<!-- ```{r apcluster} -->
<!-- corr <- round(cor(ras_data[, indep_vars],  -->
<!--                   use="pairwise.complete.obs", -->
<!--                   method = "pearson"), digits=1) -->
<!-- # order by the "Normal samples clustering -->
<!-- d1 <- as.matrix(as.dist(dist(corr), diag=T, upper=T)) -->
<!-- d1 <- 1-d1 -->
<!-- cl1 <- apcluster::apcluster(s=d1) -->
<!-- # plot correlation matrix in this order -->
<!-- neworder_names <- list() -->
<!-- for (ii in 1:length(cl1)){ -->
<!--   neworder_names[[ii]] <- names(cl1[[ii]]) -->
<!-- } -->
<!-- neworder_names <- unlist(neworder_names) -->
<!-- neworder_index <- match(neworder_names, indep_vars) -->

<!-- # plot -->
<!-- stype <- "all" -->
<!-- figfile <- file.path(cfigfolder, -->
<!--               paste("correlation_", stype, "_ap_order.pdf", sep="")) -->
<!-- p <- ggcorrplot(corr[neworder_index, neworder_index], lab=T, -->
<!--                 hc.order=F) -->
<!-- ggsave(figfile, plot=p, width=20, height =20) -->
<!-- ``` -->

  
# define ward clusters for normal and stage I only
```{r ward cluster}
ras_tmp <- ras_data[ras_data$category2 %in% c("Normal", "Tumor_Stage_I"),]
corr <- round(cor(ras_tmp[, indep_vars], 
                  use="pairwise.complete.obs",
                  method = "pearson"), digits=1)
# order by Ward method
hc <- hclust(dist(corr), method="ward.D2")
corr <- corr[hc$order, hc$order]
orig_names <- colnames(corr)
numcl <- 3
cut_clusters <- cutree(hc, k=numcl)
cut_names <- paste(as.character(cut_clusters), names(cut_clusters), sep="_")
names(cut_names) <- names(cut_clusters)
# add cluster index to names
newlabels <- cut_names[match(colnames(corr), names(cut_names))]
rownames(corr) <- newlabels
colnames(corr) <- newlabels

# plot
stype <- "Normal_and_Tumor_Stage_I"
figfile <- file.path(cfigfolder,
              paste(rtype, "_correlation_", stype, "_ward_order.pdf", sep=""))
#p <- ggcorrplot(corr[neworder_index, neworder_index], lab=T,
p <- ggcorrplot(corr, lab=T, hc.order=F)
ggsave(figfile, plot=p, width=20, height =20)

# # plot correlation matrix in this order
# neworder_names <- list()
# for (ii in 1:numcl){
#   neworder_names[[ii]] <- names(w1)[w1==ii]
# }
# neworder_index <- match(unlist(neworder_names), indep_vars)
# print("clustering of genes")
# print(neworder_names)

# # plot
# stype <- "all"
# figfile <- file.path(cfigfolder,
#               paste("correlation_", stype, "_ward_order.pdf", sep=""))
# #p <- ggcorrplot(corr[neworder_index, neworder_index], lab=T,
# p <- ggcorrplot(corr[hc$order, hc$order], lab=T,
#                 hc.order=F)
# ggsave(figfile, plot=p, width=20, height =20)
```

# PCA and outlier finding
```{r PCA}
indata_mx <- ras_data[, indep_vars]
rownames(indata_mx) <- ras_data$sample
pca <- prcomp(indata_mx, center=T, scale=T,
              tol = sqrt(.Machine$double.eps), rank.=30)
# scree plot
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

sumvar <- 0
maxi <- 1
for (pcind in 1:length(pca.var)){
  sumvar <- sumvar + pca.var.per[pcind]
  #print(c(pcind, sumvar))
  if (sumvar > 95){
    # if (sumvar >= 100){
    maxi <- pcind
    #print(maxi)
    break
  }
}
vardf <- as.data.frame(cbind('PC'=1:maxi, 'variance'=pca.var.per[1:maxi]))

p <- ggplot(data=vardf) +
  geom_col(mapping=aes(x=PC, y=cumsum(variance)))+
  xlab("Principal Component") +  ylab("Cumulative percent variation") +
  geom_hline(yintercept=95, linetype="dashed",
             color = "red", linewidth=1)
plot(p)
# save scree plot as well
figfile <- file.path(figfolder, 
                     "ras_all_scree_plot.pdf")
ggsave(figfile, plot=p, height=5, width=5)
# now plot a PCA plot with annotation
pca_res <- as_tibble(pca$x, rownames = "sample")
pca_annot <- inner_join(pca_res, ras_data, by=c("sample"))
# keep the first 6 PCs only
vardf <- vardf[1:6,]
nicelabel <- list()
for (ii in 1:nrow(vardf)){
 nicelabel[[ii]] <- paste('PC',
        as.character(ii), ' (', as.character(vardf$variance[ii]), '%)', sep="")
}

plist <- list()
for (ii in 1:(nrow(vardf)/2)){
  xname <- colnames(pca_annot)[2*ii]
  yname <- colnames(pca_annot)[2*ii+1]
  plist[[ii]] <- ggplot(data=pca_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           col  = sample_type2),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("sample type", values = c("Tumor" = "red", "Normal" = "black"))+
    labs(x = nicelabel[[2*(ii-1)+1]], y = nicelabel[[2*ii]])+
    #geom_text(mapping=aes(x = PC1, y = PC2, label=samples), size=1)+
    theme_gray(base_size = 20) # change the default font size in the grey theme  
}

pfinal <- plot_grid(plotlist=plist, ncol = 2)
resfile <- file.path(figfolder, 
                     "ras_all_pca_plot.pdf")
ggsave(resfile, plot=pfinal, width=20, height=12)

# plot the different stages as well
plist <- list()
for (ii in 1:(nrow(vardf)/2)){
  xname <- colnames(pca_annot)[2*ii]
  yname <- colnames(pca_annot)[2*ii+1]
  plist[[ii]] <- ggplot(data=pca_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           col  = category2),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category", 
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="darkblue",
                       "Tumor_Stage_II"="cyan",
                       "Tumor_Stage_III"="brown",
                       "Tumor_Stage_IV"="red",
                       "Tumor_Stage_X"="darkgreen"))+
    labs(x = nicelabel[[2*(ii-1)+1]], y = nicelabel[[2*ii]])+
    #geom_text(mapping=aes(x = PC1, y = PC2, label=samples), size=1)+
    theme_gray(base_size = 20) # change the default font size in the grey theme  
}

pfinal <- plot_grid(plotlist=plist, ncol = 2)
resfile <- file.path(figfolder, 
                     "ras_all_pca_categories.pdf")
ggsave(resfile, plot=pfinal, width=20, height=12)


```

# find potential outliers based on the pca result
```{r find outliers}
plist <- list()
outlier_list <- list()
#for (ci in bfi:bli){

outlier_list <- list()
plist <- list()
for (stype in c("Normal", "Tumor")){
  plist[[stype]] <- list()
  outlier_list[[stype]] <- c()
  for (ci in 1:nrow(vardf)){
    # assume 5 outliers, test if it's true
    normal_samples <- pca_annot[pca_annot$sample_type2=="Normal", c(1,ci+1)]
    tumor_samples <- pca_annot[pca_annot$sample_type2=="Tumor", c(1,ci+1)]

    pca_samples <- list("Normal"=normal_samples,
                        "Tumor"=tumor_samples)
    rosner_result <- rosnerTest(pca_samples[[stype]][,2,drop=T], k=5)
    otl <- rosner_result$n.outliers
    # draw quantile plot, even if there are no outliers
    #if (otl > 0){
    varname <- colnames(pca_annot)[ci+1]
    plotvector <- pca_samples[[stype]][,2,drop=T]
    names(plotvector) <- pca_samples[[stype]][,1,drop=T]
    outlier_ids <- car::qqPlot(plotvector,
                               ylab=varname, 
                               id=list(method="y", n=otl, 
                                       cex=1, col=car::carPalette()[1], location="lr"))
    plist[[stype]][[varname]] <- recordPlot()
    outlier_list[[stype]] <- c(outlier_list[[stype]], outlier_ids)
    # print(varname)
    # print(outlier_ids)
    # print("=========")
    #}
    
  }
  figfile = file.path(figfolder, 
          paste("outliers_", stype, ".pdf", sep=""))
  colfigs <- ceiling(sqrt(length(plist[[stype]])))
  p <- plot_grid(plotlist = plist[[stype]], 
                 ncol=colfigs,
                 align = 'vh', scale=0.8)
  ggsave(figfile, plot=p,  width=7*colfigs, height=5*colfigs)
}

# list outliers
outlier_all <- names(unlist(outlier_list))
outlier_all <- limma::strsplit2(outlier_all, split="\\.")[,2]

pca_annot$outlier <- "NO"
pca_annot$outlier[which(pca_annot$sample %in% outlier_all)] <- "YES"

# plot PCA with outliers
plist <- list()
for (ii in 1:(nrow(vardf)/2)){
  xname <- colnames(pca_annot)[2*ii]
  yname <- colnames(pca_annot)[2*ii+1]
  plist[[ii]] <- ggplot(data=pca_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           shape  = sample_type2, col=outlier),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("outlier", values = c("NO" = "black", "YES" = "red"))+
    scale_shape_manual("sample type", values = c("Tumor" = 19, "Normal" = 22))+
    labs(x = nicelabel[[2*(ii-1)+1]], y = nicelabel[[2*ii]])+
    #geom_text(mapping=aes(x = PC1, y = PC2, label=samples), size=1)+
    theme_gray(base_size = 20) # change the default font size in the grey theme  
}

pfinal <- plot_grid(plotlist=plist, ncol = 2)
resfile <- file.path(figfolder, 
                     "ras_all_pca_outliers.pdf")
ggsave(resfile, plot=pfinal, width=20, height=12)

# if we need the outliers
# pca_annot[pca_annot$outlier=="YES", c("sample", "outlier")]

```

# Plot LDA plot
```{r LDA normal or tumor}
indata_mx <-  as.matrix(ras_data[, indep_vars])
lda_res <- MASS::lda(indata_mx, grouping=ras_data$sample_type_factor)
lda_values <- predict(lda_res)
all.equal(lda_values$class, ras_data$sample_type_factor) # "8 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$samples <- ras_data$sample
lda_annot$sample_type_factor <- ras_data$sample_type_factor

p <- ggplot(data=lda_annot, mapping=aes(x=LD1, color=sample_type_factor))+
  geom_histogram(fill="white", position="dodge")
#facet_wrap(~sample_type2, nrow = 2)
outfig <- file.path(figfolder, "ras_all_lda_hist_tumor_or_normal.pdf")
suppressWarnings(ggsave(outfig, plot=p, width=4, height=4))
```



# Plot LDA plot for different stages as well
```{r LDA stages}
indata_mx <-  as.matrix(ras_data[, indep_vars])
ras_data$category_factor <- factor(ras_data$category2)
lda_res <- MASS::lda(indata_mx, grouping=ras_data$category_factor)
lda_values <- predict(lda_res)
all.equal(lda_values$class, ras_data$sample_type_factor) # "8 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$samples <- ras_data$sample
lda_annot$category_factor <- ras_data$category_factor

plist <- list()
for (ii in 2:ncol(lda_values$x)){
  xname <- "LD1"
  yname <- colnames(lda_annot)[ii]
  plist[[ii-1]] <- ggplot(data=lda_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           color=category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category", 
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="darkblue",
                       "Tumor_Stage_II"="cyan",
                       "Tumor_Stage_III"="brown",
                       "Tumor_Stage_IV"="red",
                       "Tumor_Stage_X"="darkgreen"))+
    theme_gray(base_size = 20) # change the default font size in the grey theme 
}

pfinal <- plot_grid(plotlist=plist, ncol = 2)
outfig <- file.path(figfolder, "ras_all_lda_categories.pdf")
ggsave(outfig, plot=pfinal, width=12, height=12)
```

```{r LDA stages but use Normal and Stages I and IV only for prediction}
ri <- which(ras_data$category2 %in% c("Normal", "Tumor_Stage_I","Tumor_Stage_IV"))
# also delete unused factors
ras_tmp <- ras_data[ri, ]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)
indata_mx <-  as.matrix(ras_tmp[, indep_vars])
lda_model <- MASS::lda(indata_mx, grouping=ras_tmp$category_factor)
lda_values <- predict(lda_model, newdata = ras_tmp[, indep_vars])
all.equal(lda_values$class, ras_tmp$category_factor) # "3 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_tmp$sample
lda_annot$category_factor <- ras_tmp$category_factor

# obtain the “proportion of trace” which is the proportion of between-class variance that is explained by successive discriminant functions. 
prop_trace <- prop.table(lda_model$svd^2)
nicelabel <- list()
for (ii in 1:length(prop_trace)){
 nicelabel[[ii]] <- paste('LD',
        as.character(ii), ' (', 
        as.character(round(100*prop_trace[ii], digits = 2)), '%)', sep="")
}

plist <- list()
for (ii in 2:ncol(lda_values$x)){
  xname <- "LD1"
  yname <- colnames(lda_annot)[ii]
  plist[[ii-1]] <- ggplot(data=lda_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           color=category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category", 
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="navy",
                       "Tumor_Stage_IV"="red"),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_IV"="Tumor Stage IV"))+
    xlab(nicelabel[[1]]) + ylab(nicelabel[[2]])+
    theme_gray(base_size = 20) # change the default font size in the grey theme 
}

pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 1)
plot(pfinal)
outfig <- file.path(figfolder, "ras_all_lda_categories_I_IV.pdf")
ggsave(outfig, plot=pfinal, width=12, height=12)

# add outliers if normal, stage I or stage IV
outlier_samples <- pca_annot$sample[pca_annot$outlier=="YES"]
print(ras_data[which(ras_data$sample %in% outlier_samples),
            c("sample", "category_factor")])
label_df <- lda_annot[which(lda_annot$sample %in% outlier_samples), ]

p <- ggplot(data=lda_annot) +
    geom_point(mapping=aes(x = LD1, y = LD2,  color=category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category",
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="navy",
                       "Tumor_Stage_IV"="red"),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_IV"="Tumor Stage IV"))+
    xlab(nicelabel[[1]]) + ylab(nicelabel[[2]])+
    theme_gray(base_size = 20)+ # change the default font size in the grey theme
  geom_text(data=label_df, mapping=aes(x = LD1, y = LD2,  color=category_factor, 
                           label=sample))
outfig <- file.path(figfolder, "ras_all_lda_categories_I_IV_plus_outliers.pdf")
ggsave(outfig, plot=p, width=12, height=12)


# plot histogram for all stages
lda_values <- predict(lda_model, newdata = ras_data[, indep_vars])
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_data$sample
lda_annot$category_factor <- ras_data$category_factor

lda_tmp <- lda_annot[!lda_annot$category_factor %in% 
                       c("Normal", "Tumor_Stage_X"),]
lda_tmp$category_factor <- droplevels(lda_tmp$category_factor)

p <- ggplot(data=lda_tmp, mapping=aes(x=LD2, color=category_factor))+
  geom_density(fill="white",  alpha=0.25)+
      scale_color_manual("category",
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="navy",
                       "Tumor_Stage_II"="purple4",
                       "Tumor_Stage_III"="purple",
                       "Tumor_Stage_IV"="red"),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_II"="Tumor Stage II",
                       "Tumor_Stage_III"="Tumor Stage III",
                       "Tumor_Stage_IV"="Tumor Stage IV"))
plot(p)
outfig <- file.path(figfolder, "ras_all_lda_density_categories.pdf")
ggsave(outfig, plot=p, width=7, height=7)

# plot LDA biplot
library(ggord)
ggord(lda_model, ras_data[, indep_vars]$category_factor)

# simply plot the varabiles
scaling_tbl <- as_tibble(lda_model$scaling, rownames="gene")
scaling_tbl <- scaling_tbl[order(scaling_tbl$LD1, decreasing = F),]
scaling_tbl$gene <- factor(scaling_tbl$gene, levels=scaling_tbl$gene)
p1 <- ggplot(scaling_tbl) +
  geom_col(aes(x=gene, y=LD1), fill = "blue", width = 0.6)+
  coord_flip()
scaling_tbl <- scaling_tbl[order(scaling_tbl$LD2, decreasing = F),]
scaling_tbl$gene <- factor(scaling_tbl$gene, levels=scaling_tbl$gene)
p2 <- ggplot(scaling_tbl) +
  geom_col(aes(x=gene, y=LD2), fill = "red", width = 0.6)+
  coord_flip()
p <- cowplot::plot_grid(p1, p2, nrow=1)
outfig <- file.path(figfolder, "lda_scaling.pdf")
ggsave(outfig, plot=p, width=10, height=6)
```

```{r LDA other visualizations for Normal Stage I and Stage IV}
ri <- which(ras_data$category2 %in% c("Normal", "Tumor_Stage_I","Tumor_Stage_IV"))
# also delete unused factors
ras_tmp <- ras_data[ri, ]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)
indata_mx <-  as.matrix(ras_tmp[, indep_vars])
lda_model <- MASS::lda(indata_mx, grouping=ras_tmp$category_factor)
lda_values <- predict(lda_model, newdata = ras_tmp[, indep_vars])
all.equal(lda_values$class, ras_tmp$category_factor) # "3 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_tmp$sample
lda_annot$category_factor <- ras_tmp$category_factor

# obtain the “proportion of trace” which is the proportion of between-class variance that is explained by successive discriminant functions. 
prop_trace <- prop.table(lda_model$svd^2)
nicelabel <- list()
for (ii in 1:length(prop_trace)){
 nicelabel[[ii]] <- paste('LD',
        as.character(ii), ' (', 
        as.character(round(100*prop_trace[ii], digits = 2)), '%)', sep="")
}

# plot violin plot for all stages
lda_values <- predict(lda_model, newdata = ras_data[, indep_vars])
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_data$sample
lda_annot$category_factor <- ras_data$category_factor

lda_tmp <- lda_annot[!lda_annot$category_factor %in% 
                       c("Normal", "Tumor_Stage_X"),]
lda_tmp$category_factor <- droplevels(lda_tmp$category_factor)

p_violin <- ggplot(lda_tmp, 
              aes(x = category_factor, y = LD2, fill=category_factor)) +
    geom_violin(trim = FALSE)+
    geom_jitter(width = 0.2, alpha = 0.6) +
    labs(x = "", y = nicelabel[[2]])+
    #labs(fill = name1)+
    scale_fill_manual("category",
            values = c("Tumor_Stage_I"="navy",
                       "Tumor_Stage_II"="purple4",
                       "Tumor_Stage_III"="purple",
                       "Tumor_Stage_IV"="red"))+
    theme(legend.position = "none")+  # Remove the legend
    coord_flip()
plot(p_violin)
outfig <- file.path(figfolder, "ras_all_lda_density_categories_violinplot.pdf")
ggsave(outfig, plot=p_violin, width=7, height=7)
```

```{r LDA stages for made up categories}
ri <- which(ras_data$category2 %in% c("Normal", "Tumor_Stage_I","Tumor_Stage_IV"))
# also delete unused factors
ras_tmp <- ras_data[ri, ]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)
# now create made up categories
set.seed(717)
ras_tmp$bogus_factor <- "black"
ti <- which(ras_tmp$category_factor %in% c("Tumor_Stage_I","Tumor_Stage_IV"))
ras_tmp$bogus_factor[ti] <- "red"
num_stageI <- length(which(ras_tmp$category_factor=="Tumor_Stage_I"))
ras_tmp$bogus_factor[sample(ti, size=num_stageI)] <- "blue"
ras_tmp$bogus_factor <- factor(ras_tmp$bogus_factor, levels=c("black", "blue", "red"))
tab_res <- table(ras_tmp$bogus_factor, ras_tmp$category_factor)
# Convert the table to a readable string
tab_text <- capture.output(print(tab_res))


indata_mx <-  as.matrix(ras_tmp[, indep_vars])
lda_model <- MASS::lda(indata_mx, grouping=ras_tmp$bogus_factor)
lda_values <- predict(lda_model, newdata = ras_tmp[, indep_vars])
all.equal(lda_values$class, ras_tmp$bogus_factor) # "3 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$bogus_factor <- ras_tmp$bogus_factor
lda_annot$category_factor <- ras_tmp$category_factor

# obtain the “proportion of trace” which is the proportion of between-class variance that is explained by successive discriminant functions. 
prop_trace <- prop.table(lda_model$svd^2)
nicelabel <- list()
for (ii in 1:length(prop_trace)){
 nicelabel[[ii]] <- paste('LD',
        as.character(ii), ' (', 
        as.character(round(100*prop_trace[ii], digits = 2)), '%)', sep="")
}

plist <- list()
for (ii in 2:ncol(lda_values$x)){
  xname <- "LD1"
  yname <- colnames(lda_annot)[ii]
  plist[[ii-1]] <- ggplot(data=lda_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           color=bogus_factor, shape=category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("bogus", 
            values = c("black" = "black",
                       "blue"="navy",
                       "red"="red"))+
    scale_shape_manual("category",
                values = c("Normal" = 1,
                       "Tumor_Stage_I"=0,
                       "Tumor_Stage_IV"=3),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_IV"="Tumor Stage IV"))+
    xlab(nicelabel[[1]]) + ylab(nicelabel[[2]])+
      annotate("text", 
           #x = Inf, y = Inf, 
           x = -4, y=4,
           label = paste(tab_text, collapse = "\n"), 
           hjust = 1, vjust = 1, size = 4) 
    theme_gray(base_size = 20) # change the default font size in the grey theme 
}

pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 1)
plot(pfinal)
outfig <- file.path(figfolder, "ras_all_lda_bogus_categories.pdf")
ggsave(outfig, plot=pfinal, width=12, height=12)

```

# repeat LDA with limited number of genes
6 best genes to separate stage I and stage IV scatter (without normal) + density plot
8 best genes (+2: best to separate normal and stage I) scatter plus density plot
```{r LDA stages but use Normal and Stages I and IV only for prediction and limited genes}
ri <- which(ras_data$category2 %in% c("Normal", "Tumor_Stage_I","Tumor_Stage_IV"))
# test_genes <- c("KRAS", "MLLT4", "SNX27", "NRAS", "HRAS")
# test_genes <- c("RASSF7", "RAPGEF6", "KRAS", "RASSF1", "NRAS", "RAPGEF2")
test_genes <- c("RGL2", "RASSF1", "RADIL", "RASSF7", "RAPGEF4", "KRAS", "RIN3",
                "PLCE1", "RAF1")
test_char <- paste0(test_genes, collapse="_")
ras_tmp <- ras_data[ri, ]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor) # also delete unused factors
indata_mx <-  as.matrix(ras_tmp[, test_genes])
lda_model <- MASS::lda(indata_mx, grouping=ras_tmp$category_factor)
lda_values <- predict(lda_model, newdata = ras_tmp[, test_genes])
all.equal(lda_values$class, ras_tmp$category_factor) # "3 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_tmp$sample
lda_annot$category_factor <- ras_tmp$category_factor

# obtain the “proportion of trace” which is the proportion of between-class variance that is explained by successive discriminant functions. 
prop_trace <- prop.table(lda_model$svd^2)
nicelabel <- list()
for (ii in 1:length(prop_trace)){
 nicelabel[[ii]] <- paste('LD',
        as.character(ii), ' (', 
        as.character(round(100*prop_trace[ii], digits = 2)), '%)', sep="")
}

plist <- list()
for (ii in 2:ncol(lda_values$x)){
  xname <- "LD1"
  yname <- colnames(lda_annot)[ii]
  plist[[ii-1]] <- ggplot(data=lda_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           color=category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category", 
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="navy",
                       "Tumor_Stage_IV"="red"),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_IV"="Tumor Stage IV"))+
    xlab(nicelabel[[1]]) + ylab(nicelabel[[2]])+
    theme_gray(base_size = 20) # change the default font size in the grey theme 
}

pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 1)
plot(pfinal)
outfig <- file.path(figfolder, 
                    paste("ras_all_lda_", test_char,".pdf"))
ggsave(outfig, plot=pfinal, width=12, height=12)

# plot histogram for all stages
lda_values <- predict(lda_model, newdata = ras_data[, test_genes])
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_data$sample
lda_annot$category_factor <- ras_data$category_factor

lda_tmp <- lda_annot[!lda_annot$category_factor %in% 
                       c("Normal", "Tumor_Stage_X"),]
lda_tmp$category_factor <- droplevels(lda_tmp$category_factor)

p <- ggplot(data=lda_tmp, mapping=aes(x=LD2, color=category_factor))+
  geom_density(fill="white",  alpha=0.25)+
      scale_color_manual("category",
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="navy",
                       "Tumor_Stage_II"="purple4",
                       "Tumor_Stage_III"="purple",
                       "Tumor_Stage_IV"="red"),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_II"="Tumor Stage II",
                       "Tumor_Stage_III"="Tumor Stage III",
                       "Tumor_Stage_IV"="Tumor Stage IV"))
plot(p)
outfig <- file.path(figfolder, 
                    paste("ras_all_lda_density",test_char,".pdf", sep=""))
ggsave(outfig, plot=p, width=7, height=7)

```

# best model finding with findAIC and BIC
```{r use stepAIC}
# now the glm using forward selection
log_min <- glm(sample_type_bool ~ 1, 
               data=ras_data, family="binomial")
log_full <- glm(as.formula(paste0("sample_type_bool ~ ", 
                                  paste(indep_vars,collapse="+"))), 
                data = ras_data, family = "binomial")
# Variance Inflation Factor (VIF): High VIF values indicate multicollinearity among predictors, 
# which can cause convergence problems.
print("variance inflation factor is huge for the full model variables")
print(car::vif(log_full))

log_best_list <- list()
# for (omethod in c("AIC", "BIC")){
for (omethod in c("AIC", "BIC")){
  if (omethod=="AIC") kval <- 2
  if (omethod=="BIC") kval <- log(nrow(ras_data))
  suppressWarnings(log_best_list[[omethod]] <- log_min %>%
    MASS::stepAIC(
      scope = list(upper = log_full, lower = ~1),
      direction="both",
      k=2, 
      trace=0))
  print(omethod)
  print(car::vif(log_best_list[[omethod]])) # now smaller vif values
}

log_best_list_normal_vs_stageI <- log_best_list

# now test the best model
set.seed(7001)
conf_mx <- runCV_LOOCV(ras_data, log_best_list[["BIC"]],
                  depvar="sample_type_factor",
                  depvar_levels=c("Normal", "Tumor"),
                  posclass="Tumor",
                  #numCV_outer=5,
                  numCV_inner=3,
                  verbose=F,
                 return_conf_mx=T)

omethod <- "BIC"
outfile <- file.path(conffolder, 
        paste(omethod, "_conf_mx_normal_vs_stageI.csv", sep=""))
print_confmx(conf_mx, outfile, coefs=log_best_list[[omethod]]$coefficients)

TClass <- factor(c(0, 0, 1, 1))
PClass <- factor(c(0, 1, 0, 1))
# y <- c(FN, TP, TN, FP)
y <- rep(0, times=4)
y[1] <- conf_mx$table[2]
y[2] <- conf_mx$table[1]
y[3] <- conf_mx$table[4]
y[4] <- conf_mx$table[3]
df <- data.frame(TClass, PClass, y)

p <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
  geom_tile(aes(fill = y), colour = "white") +
  geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_bw() + theme(legend.position = "none")

plot(p)
print(conf_mx)
outfig <- file.path(figfolder, "BIC_best_conf_mx.pdf")
ggsave(outfig, plot=p, width=7, height=7)

# # test model with only two variables
# # try with one more gene
# #test_genes <- c("PLCE1", "RGL1")
# test_genes <- c("PLCE1", "RIN1", "RAF1")
# it_model  <-  glm(as.formula(paste0("sample_type_factor ~ ",
#                                     paste(test_genes,collapse="+"))),
#                   data = ras_data, family = "binomial")
# it_aic <- it_model$aic
# conf_mx <- runCV_LOOCV(ras_data, it_model, verbose=F,
#                       return_conf_mx=T)
# 
# TClass <- factor(c(0, 0, 1, 1))
# PClass <- factor(c(0, 1, 0, 1))
# # y <- c(FN, TP, TN, FP)
# y <- rep(0, times=4)
# y[1] <- conf_mx$table[2]
# y[2] <- conf_mx$table[1]
# y[3] <- conf_mx$table[4]
# y[4] <- conf_mx$table[3]
# df <- data.frame(TClass, PClass, y)
# 
# p <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
#   geom_tile(aes(fill = y), colour = "white") +
#   geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
#   scale_fill_gradient(low = "blue", high = "red") +
#   theme_bw() + theme(legend.position = "none")
# 
# plot(p)
# print(conf_mx)
# outfig <- file.path(figfolder, "BIC_2genes_conf_mx.pdf")
# ggsave(outfig, plot=p, width=7, height=7)
# 
# it_acc <- conf_mx$byClass[["Balanced Accuracy"]]
# print(c(it_aic, it_acc, BIC(it_model)))

```


```{r use stepAIC to seperate stage I from stage IV}
# now the glm using forward selection
keep_categories <- c("Tumor_Stage_I", "Tumor_Stage_IV")
ras_tmp <- ras_data[ras_data$category2 %in% keep_categories,]
ras_tmp$category_bool  <- ifelse(ras_tmp$category2=="Tumor_Stage_IV", T, F)
log_min <- glm(category_bool ~ 1, 
               data=ras_tmp, family="binomial")
log_full <- glm(as.formula(paste0("category_bool ~ ", 
                                  paste(indep_vars,collapse="+"))), 
                data = ras_tmp, family = "binomial")
# Variance Inflation Factor (VIF): High VIF values indicate multicollinearity among predictors, 
# which can cause convergence problems.
print(car::vif(log_full))

log_best_list <- list()
for (omethod in c("AIC", "BIC")){
  if (omethod=="AIC") kval <- 2
  if (omethod=="BIC") kval <- log(nrow(ras_tmp))
  suppressWarnings(log_best_list[[omethod]] <- log_min %>%
    MASS::stepAIC(
      scope = list(upper = log_full, lower = ~1),
      direction="both",
      k=kval, 
      trace=0))
  print(omethod)
  print(car::vif(log_best_list[[omethod]])) # still huge vif values
}

# BIC(log_best_list[[2]])
# test_genes <- c("RGL2", "RASSF1","RASSF7", "RAPGEF4", "KRAS") 
# it_model  <-  glm(as.formula(paste0("category_bool ~ ",
#                                     paste(test_genes,collapse="+"))),
#                   data = ras_tmp, family = "binomial")
# BIC(it_model)

conffolder <- file.path(resfolder, "confusion_matrices")
if(!file.exists(conffolder)){dir.create(conffolder)}
# find accuracy
ras_tmp$category_factor <- factor(ras_tmp$category2,
                                 levels=c("Tumor_Stage_I", "Tumor_Stage_IV"))
# also delete unused factors
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)
set.seed(7001)
for (omethod in names(log_best_list)){
  conf_mx <- runCV_LOOCV(ras_tmp, log_best_list[[omethod]],
                  depvar="category_factor",
                  depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
                  posclass="Tumor_Stage_IV",
                  #numCV_outer=5,
                  numCV_inner=3,
                  return_conf_mx=T)
  outfile <- file.path(conffolder, 
          paste(omethod, "_conf_mx_stageI_vs_stageIV.csv", sep=""))
  print_confmx(conf_mx, outfile, coefs=log_best_list[[omethod]]$coefficients)

  TClass <- factor(c(0, 0, 1, 1))
  PClass <- factor(c(0, 1, 0, 1))
  # y <- c(FN, TP, TN, FP)
  y <- rep(0, times=4)
  y[1] <- conf_mx$table[2]
  y[2] <- conf_mx$table[1]
  y[3] <- conf_mx$table[4]
  y[4] <- conf_mx$table[3]
  df <- data.frame(TClass, PClass, y)
  
  p <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
    geom_tile(aes(fill = y), colour = "white") +
    geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
    scale_fill_gradient(low = "blue", high = "red") +
    theme_bw() + theme(legend.position = "none")
  
  plot(p)
  print(conf_mx)
  # write confusion matrix to file
  outfile <- file.path(resfolder, 
                  paste(omethod, "_best_conf_mx_categories.csv", sep=""))
  print_confmx(conf_mx, outfile)
  outfile <- gsub("_categories", "", outfile)
  print_confmx(conf_mx, outfile)
  outfig <- file.path(figfolder, 
                      paste(omethod, "_best_conf_mx_categories.pdf", sep=""))
  ggsave(outfig, plot=p, width=7, height=7)
  

  
  
  # # test with less genes
  # test_genes <- c("RGL2", "RASSF1", "RASSF7", "KRAS")
  # test_genes <- c("NRAS", "PIK3CA", "RALGDS", "RGL2", "RASSF1", "RASSF7", 
  #                 "MLLT4")
  # test_genes <- indep_vars
  # it_model  <-  glm(as.formula(paste0("category_factor ~ ",
  #                                     paste(test_genes,collapse="+"))),
  #                   data = ras_tmp, family = "binomial")
  #   conf_mx <- runCV_LOOCV(ras_tmp, it_model,
  #                 depvar="category_factor",
  #                 depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
  #                 posclass="Tumor_Stage_IV",
  #                 #numCV_outer=5,
  #                 numCV_inner=3,
  #                 return_conf_mx=T)
}
log_best_list_stageI_vs_stageIV <- log_best_list

# it_model  <-  glm(as.formula(paste0("category_factor ~ ",
#                                     paste(test_genes,collapse="+"))),
#                   data = ras_tmp, family = "binomial")
# it_aic <- it_model$aic
# conf_mx <- runCV_LOOCV(ras_tmp, it_model, 
#                   depvar="category_factor",
#                   depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
#                   posclass="Tumor_Stage_IV",
#                        verbose=F,
#                       return_conf_mx=T)
# 
# TClass <- factor(c(0, 0, 1, 1))
# PClass <- factor(c(0, 1, 0, 1))
# # y <- c(FN, TP, TN, FP)
# y <- rep(0, times=4)
# y[1] <- conf_mx$table[2]
# y[2] <- conf_mx$table[1]
# y[3] <- conf_mx$table[4]
# y[4] <- conf_mx$table[3]
# df <- data.frame(TClass, PClass, y)
# 
# p <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
#   geom_tile(aes(fill = y), colour = "white") +
#   geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
#   scale_fill_gradient(low = "blue", high = "red") +
#   theme_bw() + theme(legend.position = "none")
# 
# plot(p)
# print(conf_mx)
# outfig <- file.path(figfolder, "BIC_5genes_conf_mx_categories.pdf")
# ggsave(outfig, plot=p, width=7, height=7)
# 
# it_acc <- conf_mx$byClass[["Balanced Accuracy"]]
# print(c(it_aic, it_acc, BIC(it_model)))

```

```{r test other models}
# now the glm using forward selection
keep_categories <- c("Normal", "Tumor_Stage_I")
ras_tmp <- ras_data[ras_data$category2 %in% keep_categories,]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)
ras_tmp$category_bool  <- ifelse(ras_tmp$category2=="Tumor_Stage_I", T, F)
sel_vars <- c("RAF1", "PLCE1", "RGL1")
log_sel <- glm(as.formula(paste0("category_bool ~ ", 
                                  paste(sel_vars,collapse="+"))), 
                data = ras_tmp, family = "binomial")
# Variance Inflation Factor (VIF): High VIF values indicate multicollinearity among predictors, 
# which can cause convergence problems.
print(car::vif(log_sel))

conf_mx <- runCV(ras_tmp, log_sel,
                depvar="category_factor",
                depvar_levels=c("Normal", "Tumor_Stage_I"),
                posclass="Tumor_Stage_I",
                numCV_outer=5,
                numCV_inner=3,
                return_conf_mx=T)

print(conf_mx)

```

# Test balanced accuracy for single genes
```{r single gene balanced accuracy normal vs stageI}
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I"),
                    c(indep_vars, "category_factor")]
# also delete unused factors
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)

# calculate all measures
acc_file <- file.path(resfolder, "single_gene_model_measures_normal_vs_stageI.rds")
all_measures <- calculate_single_gene_measures(acc_file,
                                           indep_vars,
                                           ras_tmp,
                                           "category_factor",
                                           c("Normal", "Tumor_Stage_I"),
                                           "Tumor_Stage_I")

# write confusion matrix to file for selected genes
sel_genes <- c("RAF1", "PLCE1")
conffolder <- file.path(resfolder, "confusion_matrices")
if(!file.exists(conf_folder)){dir.create(conf_folder)}
for (sel_gene in sel_genes){
  it_model  <-  glm(as.formula(paste0("category_factor ~ ",
                                          paste(sel_gene,collapse="+"))),
                        data = ras_tmp, family = "binomial")
  conf_mx <- runCV_LOOCV(ras_tmp, it_model,
                  depvar="category_factor",
                  depvar_levels=c("Normal", "Tumor_Stage_I"),
                  posclass="Tumor_Stage_I",
                  numCV_inner=3,
                  return_conf_mx=T)
  outfile <- file.path(conffolder, 
                       paste("single_gene_normal_vs_stageI_", sel_gene, ".csv", sep=""))
  print_confmx(conf_mx, outfile)
}

# save to text file as well
xls_file <- gsub(".rds", ".xlsx", acc_file)
openxlsx::write.xlsx(all_measures[
  order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file)

# # now start collecting the balanced accuracy values
# acc_file <- file.path(resfolder, "balanced_accuracy_normal_vs_stageI.rds")
# if (!file.exists(acc_file)){
#   acc_list <- list()
#   for (gene in indep_vars){
#     print(gene)
#     it_model  <-  glm(as.formula(paste0("category_factor ~ ",
#                                         paste(gene,collapse="+"))),
#                       data = ras_tmp, family = "binomial")
#     acc_list[[gene]] <- runCV_LOOCV(ras_tmp, it_model, verbose=F,
#                                     depvar="category_factor",
#                      depvar_levels=c("Normal", "Tumor_Stage_I"),
#                      posclass="Tumor_Stage_I",
#                           return_conf_mx=F)
#   }  
#   saveRDS(acc_list, file=acc_file)
# } else {
#   acc_list <- readRDS(acc_file)
# }

# # create a barplot
# acc_tbl <- as.data.frame(unlist(acc_list))
# colnames(acc_tbl) <- "balanced_accuracy"
# acc_tbl <- as_tibble(acc_tbl, rownames="gene")

# put stars based on t-test
pvalue_list <- list()
for (ivar in indep_vars){
  ttest_res <- (
    t.test(ras_tmp[[ivar]][ras_tmp$category_factor=="Normal"],
           ras_tmp[[ivar]][ras_tmp$category_factor=="Tumor_Stage_I"]))
  pvalue_list[[ivar]] <- ttest_res$p.value
}
pvaldf <- as.data.frame(unlist(pvalue_list))
colnames(pvaldf) <- "p.value" # calculate adjusted p value?
pvaltbl <- as_tibble(pvaldf, rownames="gene")

# join with previous results
all_measures <- left_join(all_measures, pvaltbl, by=join_by(gene))
all_measures$ttest_stars <- stars.pval(all_measures$p.value)
all_measures <- all_measures[order(all_measures$balanced_accuracy, decreasing = F),]
all_measures$gene <- factor(all_measures$gene, levels=all_measures$gene)

# add p-values and stars then plot with the accuracy values
p <- ggplot(all_measures, aes(x = balanced_accuracy, y = gene)) +
  geom_col(fill = "gray70") +
  ## add percentage labels
  geom_text(aes(label = ttest_stars), nudge_x = 0.05) +
  geom_vline(xintercept=0.9, color="red")+
  theme_minimal()+
  xlab("Balanced accuracy") + ylab("")
outfig <- file.path(figfolder, "single_gene_balanced_acc_normal_vs_stageI.pdf")
ggsave(outfig, plot=p, width=4, height=7)
```

```{r single gene balanced accuracy normal vs tumor}
ras_tmp <- ras_data[ras_data$sample_type2 %in% 
                      c("Normal", "Tumor"),
                    c(indep_vars, "sample_type_factor")]
# also delete unused factors
ras_tmp$sample_type_factor <- droplevels(ras_tmp$sample_type_factor)

# calculate all measures
acc_file <- file.path(resfolder, "single_gene_model_measures_normal_vs_tumor.rds")
all_measures <- calculate_single_gene_measures(acc_file,
                                           indep_vars,
                                           ras_tmp,
                                           "sample_type_factor",
                                           c("Normal", "Tumor"),
                                           "Tumor")

# write confusion matrix to file for selected genes
sel_genes <- c("PLCE1")
conffolder <- file.path(resfolder, "confusion_matrices")
if(!file.exists(conf_folder)){dir.create(conf_folder)}
for (sel_gene in sel_genes){
  it_model  <-  glm(as.formula(paste0("sample_type_factor ~ ",
                                          paste(sel_gene,collapse="+"))),
                        data = ras_tmp, family = "binomial")
  conf_mx <- runCV_LOOCV(ras_tmp, it_model,
                  depvar="sample_type_factor",
                  depvar_levels=c("Normal", "Tumor"),
                  posclass="Tumor",
                  numCV_inner=3,
                  return_conf_mx=T)
  outfile <- file.path(conffolder, 
                       paste("single_gene_normal_vs_tumor_", sel_gene, ".csv", sep=""))
  print_confmx(conf_mx, outfile)
}
  

# save to text file as well
xls_file <- gsub(".rds", ".xlsx", acc_file)
openxlsx::write.xlsx(all_measures[
  order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file)

# put stars based on t-test
pvalue_list <- list()
for (ivar in indep_vars){
  ttest_res <- (
    t.test(ras_tmp[[ivar]][ras_tmp$sample_type_factor=="Normal"],
           ras_tmp[[ivar]][ras_tmp$sample_type_factor=="Tumor"]))
  pvalue_list[[ivar]] <- ttest_res$p.value
}
pvaldf <- as.data.frame(unlist(pvalue_list))
colnames(pvaldf) <- "p.value" # calculate adjusted p value?
pvaltbl <- as_tibble(pvaldf, rownames="gene")

all_measures <- left_join(all_measures, pvaltbl, by=join_by(gene))
all_measures$ttest_stars <- stars.pval(all_measures$p.value)
all_measures <- all_measures[order(all_measures$balanced_accuracy, decreasing = F),]
all_measures$gene <- factor(all_measures$gene, levels=all_measures$gene)

# add p-values and stars then plot with the accuracy values
p <- ggplot(all_measures, aes(x = balanced_accuracy, y = gene)) +
  geom_col(fill = "gray70") +
  ## add percentage labels
  geom_text(aes(label = ttest_stars), nudge_x = 0.05) +
  geom_vline(xintercept=0.9, color="red")+
  theme_minimal()+
  xlab("Balanced accuracy") + ylab("")
outfig <- file.path(figfolder, "single_gene_balanced_acc_normal_vs_tumor.pdf")
ggsave(outfig, plot=p, width=4, height=7)
```

```{r single gene balanced accuracy stage I vs stage IV}
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(indep_vars, "category_factor")]
# also delete unused factors
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)

# calculate all measures
acc_file <- file.path(resfolder, "single_gene_model_measures_stageI_vs_stageVI.rds")
all_measures <- calculate_single_gene_measures(acc_file,
                                           indep_vars,
                                           ras_tmp,
                                           "category_factor",
                                           c("Tumor_Stage_I", "Tumor_Stage_IV"),
                                           "Tumor_Stage_IV")
# save to text file as well
xls_file <- gsub(".rds", ".xlsx", acc_file)
openxlsx::write.xlsx(all_measures[
  order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file)

# put stars based on t-test
pvalue_list <- list()
for (ivar in indep_vars){
  ttest_res <- (
    t.test(ras_tmp[[ivar]][ras_tmp$category_factor=="Tumor_Stage_I"],
           ras_tmp[[ivar]][ras_tmp$category_factor=="Tumor_Stage_IV"]))
  pvalue_list[[ivar]] <- ttest_res$p.value
}
pvaldf <- as.data.frame(unlist(pvalue_list))
colnames(pvaldf) <- "p.value" # calculate adjusted p value?
pvaltbl <- as_tibble(pvaldf, rownames="gene")

all_measures <- left_join(all_measures, pvaltbl, by=join_by(gene))
all_measures$ttest_stars <- stars.pval(all_measures$p.value)
all_measures <- all_measures[order(all_measures$balanced_accuracy, decreasing = F),]
all_measures$gene <- factor(all_measures$gene, levels=all_measures$gene)

# add p-values and stars then plot with the accuracy values
p <- ggplot(all_measures, aes(x = balanced_accuracy, y = gene)) +
  geom_col(fill = "gray70") +
  ## add percentage labels
  geom_text(aes(label = ttest_stars), nudge_x = 0.05) +
  #geom_vline(xintercept=0.9, color="red")+
  theme_minimal()+
  xlab("Balanced accuracy") + ylab("")
outfig <- file.path(figfolder, "single_gene_balanced_acc_stageI_vs_stageIV.pdf")
ggsave(outfig, plot=p, width=4, height=7)

```

<!-- # Test balanced accuracy for pair of genes -->
<!-- ```{r single gene balanced accuracy normal vs stageI} -->
<!-- ras_tmp <- ras_data[ras_data$category2 %in%  -->
<!--                       c("Normal", "Tumor_Stage_I"), -->
<!--                     c(indep_vars, "category_factor")] -->
<!-- # also delete unused factors -->
<!-- ras_tmp$category_factor <- droplevels(ras_tmp$category_factor) -->

<!-- # calculate all measures -->
<!-- acc_file <- file.path(resfolder, "double_gene_model_measures_normal_vs_stageI.rds") -->
<!-- all_measures <- calculate_double_gene_measures(acc_file, -->
<!--                                            indep_vars, -->
<!--                                            ras_tmp, -->
<!--                                            "category_factor", -->
<!--                                            c("Normal", "Tumor_Stage_I"), -->
<!--                                            "Tumor_Stage_I") -->
<!-- # save to text file as well -->
<!-- xls_file <- gsub(".rds", ".xlsx", acc_file) -->
<!-- openxlsx::write.xlsx(all_measures[ -->
<!--   order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file) -->

<!-- # plot gene pairs with accuracy above 0.95 -->
<!-- indata <- all_measures[which(all_measures$balanced_accuracy>=0.95),] -->
<!-- # order for plot -->
<!-- indata <- indata[order(indata$balanced_accuracy, decreasing = F),] -->
<!-- indata$gene_pair <- factor(indata$gene_pair, levels=indata$gene_pair) -->
<!-- p <- ggplot(indata, aes(x = balanced_accuracy, y = gene_pair)) + -->
<!--   geom_col(fill = "gray70") + -->
<!--   ## add percentage labels -->
<!--   geom_text(aes(label = gene_pair, x=0.2), nudge_x = -0.05) + -->
<!--   # geom_vline(xintercept=0.95, color="red")+ -->
<!--   theme(axis.title.y=element_blank(), -->
<!--         axis.text.y=element_blank(), -->
<!--         axis.ticks.y=element_blank())+ -->
<!--   xlab("Balanced accuracy")  -->
<!-- outfig <- file.path(figfolder, "double_gene_balanced_acc_normal_vs_stageI_thr095.pdf") -->
<!-- ggsave(outfig, plot=p, width=4, height=7) -->
<!-- ``` -->

<!-- # Plot balanced accuracy pair of genes - should reproduce previous plot -->
<!-- ```{r two genes balanced accuracy normal vs stageI} -->
<!-- excelfile <- file.path(phome, "Results", -->
<!--           "archive_new/best_model_2genes_normal_vs_stageI.xlsx") -->
<!-- indata <- read_excel(excelfile, sheet=rtype) -->
<!-- indata <- indata[indata$gene1 %in% indep_vars & indata$gene2 %in% indep_vars,] -->
<!-- indata$index <- 1:nrow(indata) -->
<!-- indata <- indata[indata$acc>=0.95,] -->
<!-- indata <- indata[order(indata$acc, decreasing = F),] -->
<!-- indata$label <- paste(indata$gene1, indata$gene2, sep=" + ") -->
<!-- indata$label <- factor(indata$label, levels=indata$label) -->
<!-- # plot gene pairs with accuracy above 0.9 -->
<!-- p <- ggplot(indata, aes(x = acc, y = label)) + -->
<!--   geom_col(fill = "gray70") + -->
<!--   ## add percentage labels -->
<!--   geom_text(aes(label = label, x=0.2), nudge_x = -0.05) + -->
<!--   # geom_vline(xintercept=0.95, color="red")+ -->
<!--   theme(axis.title.y=element_blank(), -->
<!--         axis.text.y=element_blank(), -->
<!--         axis.ticks.y=element_blank())+ -->
<!--   xlab("Balanced accuracy")  -->
<!-- outfig <- file.path(figfolder, "two_genes_balanced_acc_normal_vs_stageI_thr095.pdf") -->
<!-- ggsave(outfig, plot=p, width=7, height=10) -->

<!-- # additionally plot scatter plot for the pair of genes -->
<!-- ras_tmp <- ras_data[ras_data$category2 %in%  -->
<!--                       c("Normal", "Tumor_Stage_I"), -->
<!--                     c(indep_vars, "category_factor", "category2")] -->
<!-- # also delete unused factors -->
<!-- ras_tmp$category_factor <- droplevels(ras_tmp$category_factor) -->
<!-- plist <- list() -->
<!-- for (ii in 1:nrow(indata)){ -->
<!--   gene1 <- indata$gene1[[ii]] -->
<!--   gene2 <- indata$gene2[[ii]] -->
<!--   plist[[ii]] <- ggplot(data=ras_tmp,  -->
<!--                         mapping=aes(x=.data[[gene1]], y=.data[[gene2]], -->
<!--                         col=category2))+ -->
<!--     geom_point(shape=1)+ -->
<!--     scale_color_manual("category", -->
<!--             values = c("Tumor_Stage_I" = "darkblue", -->
<!--                        "Normal"="black"), -->
<!--             labels = c("Tumor_Stage_I" = "Stage I", -->
<!--                        "Normal"="Normal")) -->
<!-- } -->
<!-- plist <- plist[length(plist):1] -->
<!-- pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 2) -->
<!-- outfig <- file.path(figfolder, -->
<!--     "two_genes_balanced_acc_normal_vs_stageI_thr095_scatterplots.pdf") -->
<!-- ggsave(outfig, plot=pfinal, width = 7, height=35) -->

<!-- sel_genes <- c("RAF1+RGL1", "RAF1+PIK3CG", "RAF1+PLCE1", "RAF1+APBB1IP", "KRAS+RAF1", -->
<!-- "KRAS+RGL1", "RAF1+PIK3CD", "RAF1+RASSF5", "RAF1+RADIL", "RGL1+PLCE1",  -->
<!-- "PLCE1+RIN1", "PLCE1+RADIL", "PLCE1+MYO10", "NRAS+RAF1", "RAF1+RGL2",  -->
<!-- "RAF1+RIN3", "RAF1+KRIT1", "RAF1+RGS14", "PLCE1+KRIT1", "PLCE1+MLLT4", -->
<!-- "ARAF+RIN1", "RGL1+RGL2", "BRAF+RAF1", "RAF1+PIK3C2G", "RAF1+RIN2", -->
<!-- "RAF1+ARAP3", "RAF1+RASSF7", "RAF1+RAPGEF4", "RAF1+RAPH1", "RAF1+GRB7", -->
<!-- "RAF1+MLLT4", "PIK3C2B+PLCE1", "PIK3C2G+PLCE1", "PLCE1+GRB7", "RIN1+KRIT1", -->
<!-- "HRAS+PLCE1", "PIK3CB+PLCE1", "PLCE1+RAPGEF6") -->
<!-- conffolder <- file.path(resfolder, "confusion_matrices") -->
<!-- if(!file.exists(conf_folder)){dir.create(conf_folder)} -->
<!-- for (sel_gene in sel_genes){ -->
<!--   it_model  <-  glm(as.formula(paste0("category_factor ~ ", -->
<!--                                           sel_gene)), -->
<!--                         data = ras_tmp, family = "binomial") -->
<!--   conf_mx <- runCV_LOOCV(ras_tmp, it_model, -->
<!--                   depvar="category_factor", -->
<!--                   depvar_levels=c("Normal", "Tumor_Stage_I"), -->
<!--                   posclass="Tumor_Stage_I", -->
<!--                   numCV_inner=3, -->
<!--                   return_conf_mx=T) -->
<!--   outfile <- file.path(conffolder,  -->
<!--                 paste("double_genes_normal_vs_stageI_", sel_gene, ".csv", sep="")) -->
<!--   print_confmx(conf_mx, outfile) -->
<!-- } -->

<!-- ``` -->

<!-- ```{r two genes balanced accuracy stageI vs stageIV} -->
<!-- excelfile <- file.path(phome, "Results", -->
<!--           "archive_new/best_model_2genes_stageI_vs_stageIV.xlsx") -->
<!-- indata <- read_excel(excelfile, sheet=rtype) -->
<!-- indata <- indata[indata$gene1 %in% indep_vars & indata$gene2 %in% indep_vars,] -->
<!-- indata$index <- 1:nrow(indata) -->
<!-- indata <- indata[indata$acc>=0.65,] -->
<!-- indata <- indata[order(indata$acc, decreasing = F),] -->
<!-- indata$label <- paste(indata$gene1, indata$gene2, sep=" + ") -->
<!-- indata$label <- factor(indata$label, levels=indata$label) -->
<!-- # plot gene pairs with accuracy above 0.9 -->
<!-- p <- ggplot(indata, aes(x = acc, y = label)) + -->
<!--   geom_col(fill = "gray70") + -->
<!--   ## add percentage labels -->
<!--   geom_text(aes(label = label, x=0.2), nudge_x = -0.05) + -->
<!--   #geom_vline(xintercept=0.95, color="red")+ -->
<!--   theme(axis.title.y=element_blank(), -->
<!--         axis.text.y=element_blank(), -->
<!--         axis.ticks.y=element_blank())+ -->
<!--   xlab("Balanced accuracy") -->
<!-- outfig <- file.path(figfolder, "two_genes_balanced_acc_stageI_vs_stageIV_thr065.pdf") -->
<!-- ggsave(outfig, plot=p, width=7, height=7) -->

<!-- # additionally plot scatter plot for the pair of genes -->
<!-- ras_tmp <- ras_data[ras_data$category2 %in%  -->
<!--                       c("Tumor_Stage_I", "Tumor_Stage_IV"), -->
<!--                     c(indep_vars, "category_factor", "category2")] -->
<!-- # also delete unused factors -->
<!-- ras_tmp$category_factor <- droplevels(ras_tmp$category_factor) -->
<!-- plist <- list() -->
<!-- for (ii in 1:nrow(indata)){ -->
<!--   gene1 <- indata$gene1[[ii]] -->
<!--   gene2 <- indata$gene2[[ii]] -->
<!--   plist[[ii]] <- ggplot(data=ras_tmp,  -->
<!--                         mapping=aes(x=.data[[gene1]], y=.data[[gene2]], -->
<!--                         col=category2))+ -->
<!--     geom_point(shape=1)+ -->
<!--     scale_color_manual("category", -->
<!--             values = c("Tumor_Stage_I" = "darkblue", -->
<!--                        "Tumor_Stage_IV"="red"), -->
<!--             labels = c("Tumor_Stage_I" = "Stage I", -->
<!--                        "Tumor_Stage_IV"="Stage IV")) -->
<!-- } -->
<!-- plist <- plist[length(plist):1] -->
<!-- pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 2) -->
<!-- outfig <- file.path(figfolder, -->
<!--     "two_genes_balanced_acc_stageI_vs_stageIV_thr065_scatterplots.pdf") -->
<!-- ggsave(outfig, plot=pfinal, width = 7, height=21) -->
<!-- ``` -->

<!-- Calculate for normal vs tumor which was not done previously -->
<!-- ```{r double gene balanced accuracy normal vs tumor} -->
<!-- ras_tmp <- ras_data[ras_data$sample_type2 %in%  -->
<!--                       c("Normal", "Tumor"), -->
<!--                     c(indep_vars, "sample_type_factor")] -->
<!-- # also delete unused factors -->
<!-- ras_tmp$sample_type_factor <- droplevels(ras_tmp$sample_type_factor) -->

<!-- # calculate all measures -->
<!-- acc_file <- file.path(resfolder, "two_gene_model_measures_normal_vs_tumor.rds") -->
<!-- all_measures <- calculate_double_gene_measures(acc_file, -->
<!--                                            indep_vars, -->
<!--                                            ras_tmp, -->
<!--                                            "sample_type_factor", -->
<!--                                            c("Normal", "Tumor"), -->
<!--                                            "Tumor") -->
<!-- # save to text file as well -->
<!-- xls_file <- gsub(".rds", ".xlsx", acc_file) -->
<!-- # change '.' to '+' -->
<!-- all_measures$gene_pair <- gsub("\\.", "+", all_measures$gene_pair) -->
<!-- all_measures$gene1 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,1] -->
<!-- all_measures$gene2 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,2] -->
<!-- openxlsx::write.xlsx(all_measures[ -->
<!--   order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file) -->

<!-- # write confusion matrix to file for selected genes -->
<!-- sel_genes <- c("RGL1+PLCE1",  -->
<!-- "PLCE1+MYO10", "PIK3C2B+PLCE1", "RAF1+PLCE1", "PLCE1+RADIL", "PLCE1+GRB7") -->
<!-- conffolder <- file.path(resfolder, "confusion_matrices") -->
<!-- if(!file.exists(conf_folder)){dir.create(conf_folder)} -->
<!-- for (sel_gene in sel_genes){ -->
<!--   it_model  <-  glm(as.formula(paste0("sample_type_factor ~ ", -->
<!--                                           sel_gene)), -->
<!--                         data = ras_tmp, family = "binomial") -->
<!--   conf_mx <- runCV_LOOCV(ras_tmp, it_model, -->
<!--                   depvar="sample_type_factor", -->
<!--                   depvar_levels=c("Normal", "Tumor"), -->
<!--                   posclass="Tumor", -->
<!--                   numCV_inner=3, -->
<!--                   return_conf_mx=T) -->
<!--   outfile <- file.path(conffolder,  -->
<!--                 paste("double_genes_normal_vs_tumor_", sel_gene, ".csv", sep="")) -->
<!--   print_confmx(conf_mx, outfile) -->
<!-- } -->


<!-- # plot gene pairs with accuracy above 0.95 -->
<!-- indata <- all_measures[which(all_measures$balanced_accuracy>=0.9),] -->
<!-- # order for plot -->
<!-- indata <- indata[order(indata$balanced_accuracy, decreasing = F),] -->
<!-- indata$gene_pair <- factor(indata$gene_pair, levels=indata$gene_pair) -->

<!-- p <- ggplot(indata, aes(x = balanced_accuracy, y = gene_pair)) + -->
<!--   geom_col(fill = "gray70") + -->
<!--   ## add percentage labels -->
<!--   geom_text(aes(label = gene_pair, x=0.2), nudge_x = -0.05) + -->
<!--   # geom_vline(xintercept=0.95, color="red")+ -->
<!--   theme(axis.title.y=element_blank(), -->
<!--         axis.text.y=element_blank(), -->
<!--         axis.ticks.y=element_blank())+ -->
<!--   xlab("Balanced accuracy") -->
<!-- outfig <- file.path(figfolder, "double_gene_balanced_acc_normal_vs_tumor_thr09.pdf") -->
<!-- ggsave(outfig, plot=p, width = 7, height=14) -->
<!-- ``` -->



# Plot balanced accuracy pair of genes 
```{r double gene balanced accuracy normal vs stageI}
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I"),
                    c(indep_vars, "category_factor")]
# also delete unused factors
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)

# calculate all measures
acc_file <- file.path(resfolder, "two_gene_model_measures_normal_vs_stageI.rds")
all_measures <- calculate_double_gene_measures(acc_file,
                                           indep_vars,
                                           ras_tmp,
                                           "category_factor",
                                           c("Normal", "Tumor_Stage_I"),
                                           "Tumor_Stage_I")
# save to text file as well
xls_file <- gsub(".rds", ".xlsx", acc_file)
# change '.' to '+'
all_measures$gene_pair <- gsub("\\.", "+", all_measures$gene_pair)
all_measures$gene1 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,1]
all_measures$gene2 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,2]
openxlsx::write.xlsx(all_measures[
  order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file)

# # write confusion matrix to file for selected genes
# sel_genes <- c("RGL1+PLCE1", 
# "PLCE1+MYO10", "PIK3C2B+PLCE1", "RAF1+PLCE1", "PLCE1+RADIL", "PLCE1+GRB7")
# conffolder <- file.path(resfolder, "confusion_matrices")
# if(!file.exists(conf_folder)){dir.create(conf_folder)}
# for (sel_gene in sel_genes){
#   it_model  <-  glm(as.formula(paste0("category_factor ~ ",
#                                           sel_gene)),
#                         data = ras_tmp, family = "binomial")
#   conf_mx <- runCV_LOOCV(ras_tmp, it_model,
#                   depvar="category_factor",
#                   depvar_levels=c("Normal", "Tumor_Stage_I"),
#                   posclass="Tumor_Stage_I",
#                   numCV_inner=3,
#                   return_conf_mx=T)
#   outfile <- file.path(conffolder, 
#                 paste("double_genes_normal_vs_tumor_", sel_gene, ".csv", sep=""))
#   print_confmx(conf_mx, outfile)
# }
  

# plot gene pairs with accuracy above 0.90
indata <- all_measures[which(all_measures$balanced_accuracy>=0.9),]
# order for plot
indata <- indata[order(indata$balanced_accuracy, decreasing = F),]
indata$gene_pair <- factor(indata$gene_pair, levels=indata$gene_pair)

p <- ggplot(indata, aes(x = balanced_accuracy, y = gene_pair)) +
  geom_col(fill = "gray70") +
  ## add percentage labels
  geom_text(aes(label = gene_pair, x=0.2), nudge_x = -0.05) +
  # geom_vline(xintercept=0.95, color="red")+
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())+
  xlab("Balanced accuracy")
outfig <- file.path(figfolder, "double_gene_balanced_acc_normal_vs_stageI_thr09.pdf")
ggsave(outfig, plot=p, width = 7, height=14)
```

```{r double gene balanced accuracy stageI vs stageIV}
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(indep_vars, "category_factor")]
# also delete unused factors
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)

# calculate all measures
acc_file <- file.path(resfolder, "two_gene_model_measures_stageI_vs_stageIV.rds")
all_measures <- calculate_double_gene_measures(acc_file,
                                           indep_vars,
                                           ras_tmp,
                                           "category_factor",
                                           c("Tumor_Stage_I", "Tumor_Stage_IV"),
                                           "Tumor_Stage_IV")
# save to text file as well
xls_file <- gsub(".rds", ".xlsx", acc_file)
# change '.' to '+'
all_measures$gene_pair <- gsub("\\.", "+", all_measures$gene_pair)
all_measures$gene1 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,1]
all_measures$gene2 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,2]
openxlsx::write.xlsx(all_measures[
  order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file)

# # write confusion matrix to file for selected genes
# sel_genes <- c("RGL1+PLCE1", 
# "PLCE1+MYO10", "PIK3C2B+PLCE1", "RAF1+PLCE1", "PLCE1+RADIL", "PLCE1+GRB7")
# conffolder <- file.path(resfolder, "confusion_matrices")
# if(!file.exists(conf_folder)){dir.create(conf_folder)}
# for (sel_gene in sel_genes){
#   it_model  <-  glm(as.formula(paste0("category_factor ~ ",
#                                           sel_gene)),
#                         data = ras_tmp, family = "binomial")
#   conf_mx <- runCV_LOOCV(ras_tmp, it_model,
#                   depvar="category_factor",
#                   depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
#                   posclass="Tumor_Stage_IV",
#                   numCV_inner=3,
#                   return_conf_mx=T)
#   outfile <- file.path(conffolder, 
#                 paste("double_genes_stageI_vs_stageIV_", sel_gene, ".csv", sep=""))
#   print_confmx(conf_mx, outfile)
# }
  

# plot gene pairs with accuracy above 0.60
indata <- all_measures[which(all_measures$balanced_accuracy>=0.6),]
# order for plot
indata <- indata[order(indata$balanced_accuracy, decreasing = F),]
indata$gene_pair <- factor(indata$gene_pair, levels=indata$gene_pair)

p <- ggplot(indata, aes(x = balanced_accuracy, y = gene_pair)) +
  geom_col(fill = "gray70") +
  ## add percentage labels
  geom_text(aes(label = gene_pair, x=0.2), nudge_x = -0.05) +
  # geom_vline(xintercept=0.95, color="red")+
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())+
  xlab("Balanced accuracy")
outfig <- file.path(figfolder, "double_gene_balanced_acc_stageI_vs_stageIV_thr06.pdf")
ggsave(outfig, plot=p, width = 7, height=14)
```

```{r double gene balanced accuracy normal vs tumor}
ras_tmp <- ras_data[ras_data$sample_type2 %in% 
                      c("Normal", "Tumor"),
                    c(indep_vars, "sample_type_factor")]
# also delete unused factors
ras_tmp$sample_type_factor <- droplevels(ras_tmp$sample_type_factor)

# calculate all measures
acc_file <- file.path(resfolder, "two_gene_model_measures_normal_vs_tumor.rds")
all_measures <- calculate_double_gene_measures(acc_file,
                                           indep_vars,
                                           ras_tmp,
                                           "sample_type_factor",
                                           c("Normal", "Tumor"),
                                           "Tumor")
# save to text file as well
xls_file <- gsub(".rds", ".xlsx", acc_file)
# change '.' to '+'
all_measures$gene_pair <- gsub("\\.", "+", all_measures$gene_pair)
all_measures$gene1 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,1]
all_measures$gene2 <- limma::strsplit2(all_measures$gene_pair, split="\\+")[,2]
openxlsx::write.xlsx(all_measures[
  order(all_measures$balanced_accuracy, decreasing = T),], file=xls_file)

# # write confusion matrix to file for selected genes
# sel_genes <- c("RGL1+PLCE1", 
# "PLCE1+MYO10", "PIK3C2B+PLCE1", "RAF1+PLCE1", "PLCE1+RADIL", "PLCE1+GRB7")
# conffolder <- file.path(resfolder, "confusion_matrices")
# if(!file.exists(conf_folder)){dir.create(conf_folder)}
# for (sel_gene in sel_genes){
#   it_model  <-  glm(as.formula(paste0("sample_type_factor ~ ",
#                                           sel_gene)),
#                         data = ras_tmp, family = "binomial")
#   conf_mx <- runCV_LOOCV(ras_tmp, it_model,
#                   depvar="sample_type_factor",
#                   depvar_levels=c("Normal", "Tumor"),
#                   posclass="Tumor",
#                   numCV_inner=3,
#                   return_conf_mx=T)
#   outfile <- file.path(conffolder, 
#                 paste("double_genes_normal_vs_tumor_", sel_gene, ".csv", sep=""))
#   print_confmx(conf_mx, outfile)
# }
  

# plot gene pairs with accuracy above 0.95
indata <- all_measures[which(all_measures$balanced_accuracy>=0.9),]
# order for plot
indata <- indata[order(indata$balanced_accuracy, decreasing = F),]
indata$gene_pair <- factor(indata$gene_pair, levels=indata$gene_pair)

p <- ggplot(indata, aes(x = balanced_accuracy, y = gene_pair)) +
  geom_col(fill = "gray70") +
  ## add percentage labels
  geom_text(aes(label = gene_pair, x=0.2), nudge_x = -0.05) +
  # geom_vline(xintercept=0.95, color="red")+
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())+
  xlab("Balanced accuracy")
outfig <- file.path(figfolder, "double_gene_balanced_acc_normal_vs_tumor_thr09.pdf")
ggsave(outfig, plot=p, width = 7, height=14)
```


# use ANOVA FOR THE STAGES
```{r calculate anova for stages}
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I", "Tumor_Stage_IV"),]
plotlist <- list()
for (gene in indep_vars){
  stagelm <- lm(as.formula(paste0(gene, " ~ category_factor")), data=ras_tmp)
  stageav <- aov(stagelm)
  summary(stageav)
  tukeydf <- as_tibble(TukeyHSD(stageav, conf.level=.95)$category_factor, 
                       rownames="comparison")
  tukeydf$p.signif <- stars.pval(tukeydf$`p adj`)
  tukeydf$p.signif[tukeydf$p.signif==" "] <- "ns"
  tukeydf$group1 <- limma::strsplit2(tukeydf$comparison, split="-")[,2]
  tukeydf$group2 <- limma::strsplit2(tukeydf$comparison, split="-")[,1]
  
  # use Tukey later, but test with the example given
  # tmpdf <- compare_means(as.formula(paste0(gene, " ~ category_factor")), 
  #                        data=ras_tmp)
  my_comparisons <- list( c("Normal", "Tumor_Stage_I"), 
                          c("Normal", "Tumor_Stage_IV"), 
                          c("Tumor_Stage_I", "Tumor_Stage_IV") )
  # set y positions on top of highest notch
  p1 <- ggplot(data=ras_tmp)+
    geom_boxplot(mapping=aes(x=category_factor, y=.data[[gene]]))+
    scale_x_discrete(labels=c("Normal","Stage I","Stage IV"))+
    theme(axis.title.x=element_blank(),
          axis.text.x = element_text(angle = 45, vjust = 0.4, hjust=1),
          plot.margin = unit(c(0.05, 0.05, 0.3, 0.05), 
                                "inches"))
  ymax <- median(ras_tmp[[gene]]) + 2 * IQR(ras_tmp[[gene]])
  ymin <- median(ras_tmp[[gene]]) - 2 * IQR(ras_tmp[[gene]])
  ystep <- (ymax-ymin)/10
  p2 <- p1+ stat_pvalue_manual(tukeydf, label = "p.signif", 
    y.position = c(ymax+0.1*ystep, ymax+1.1*ystep, ymax+2.1*ystep))
  plotlist[[gene]] <- p2
}
# create a plot with normal, tumor and stage I
plot_all <- cowplot::plot_grid(plotlist=plotlist,
                                nrow=ceiling(sqrt(length(plotlist))))
figfile <- file.path(figfolder, "tukey_grid.pdf")
ggsave(figfile, plot=plot_all, width=20, height=20, limitsize = F)

```

# Test how well a much smaller amount of genes could seperate 
normal from tumor stage I 
```{r lower sample size}
gene_list <- c("PLCE1", "RGL1", "RAF1")
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(gene_list, "category_factor")]
# now select the given amount of samples
# option  1) 8 normal, 5 stage I, 10 stage IV
# option 2) 5 normal, 10 stage I, 6 stage IV
# randomly select these number of samples
normal_index <- which(ras_tmp$category_factor=="Normal")
normal_selected <- sample(normal_index, size=5)
stageI_index <- which(ras_tmp$category_factor=="Tumor_Stage_I")
stageI_selected <- sample(stageI_index, size=10)
stageIV_index <- which(ras_tmp$category_factor=="Tumor_Stage_IV")
stageIV_selected <- sample(stageIV_index, size=6)

ras_tmp2 <- ras_tmp[c(normal_selected, stageI_selected, stageIV_selected),]
# scale the new dataframe
ras_tmp2[, gene_list] <- scale(ras_tmp2[, gene_list])
# repeat the anova plots
plotlist <- list()
for (gene in gene_list){
  stagelm <- lm(as.formula(paste0(gene, " ~ category_factor")), data=ras_tmp2)
  stageav <- aov(stagelm)
  summary(stageav)
  tukeydf <- as_tibble(TukeyHSD(stageav, conf.level=.95)$category_factor, 
                       rownames="comparison")
  tukeydf$p.signif <- stars.pval(tukeydf$`p adj`)
  tukeydf$p.signif[tukeydf$p.signif==" "] <- "ns"
  tukeydf$group1 <- limma::strsplit2(tukeydf$comparison, split="-")[,2]
  tukeydf$group2 <- limma::strsplit2(tukeydf$comparison, split="-")[,1]
  
  # use Tukey later, but test with the example given
  tmpdf <- compare_means(as.formula(paste0(gene, " ~ category_factor")), data=ras_tmp)
  my_comparisons <- list( c("Normal", "Tumor_Stage_I"), 
                          c("Normal", "Tumor_Stage_IV"), 
                          c("Tumor_Stage_I", "Tumor_Stage_IV") )
  # set y positions on top of highest notch
  p1 <- ggplot(data=ras_tmp2)+
    geom_boxplot(mapping=aes(x=category_factor, y=.data[[gene]]))+
    scale_x_discrete(labels=c("Normal","Stage I","Stage IV"))+
    theme(axis.title.x=element_blank(),
          axis.text.x = element_text(angle = 45, vjust = 0.4, hjust=1),
          plot.margin = unit(c(0.05, 0.05, 0.3, 0.05), 
                                "inches"))
  ymax <- median(ras_tmp2[[gene]]) + 2 * IQR(ras_tmp2[[gene]])
  ymin <- median(ras_tmp2[[gene]]) - 2 * IQR(ras_tmp2[[gene]])
  ystep <- (ymax-ymin)/10
  p2 <- p1+ stat_pvalue_manual(tukeydf, label = "p.signif", 
    y.position = c(ymax+0.1*ystep, ymax+1.1*ystep, ymax+2.1*ystep))
  plotlist[[gene]] <- p2
}
# create a plot with normal, tumor and stage I
plot_all <- cowplot::plot_grid(plotlist=plotlist, nrow=1)
figfile <- file.path(figfolder, "best_genes_reduced_tukey_normal_vs_stageI.pdf")
ggsave(figfile, plot=plot_all, width=14, height=5)

# can we seperate normal from tumor stage I using these genes?
ras_tmp3 <- ras_tmp2[ras_tmp2$category_factor %in% c("Normal", "Tumor_Stage_I"),]
ras_tmp3$category_factor <- droplevels(ras_tmp3$category_factor)
it_model  <-  glm(as.formula(paste0("category_factor ~ ",
                                    paste(gene_list,collapse="+"))),
                  data = ras_data, family = "binomial")
conf_mx <- runCV(ras_tmp3, it_model,
                 depvar="category_factor",
                 depvar_levels=c("Normal", "Tumor_Stage_I"),
                 posclass="Tumor_Stage_I",
                 return_conf_mx=T)
#print(conf_mx)
# plot conf_mx
TClass <- factor(c(0, 0, 1, 1))
PClass <- factor(c(0, 1, 0, 1))
# y <- c(FN, TP, TN, FP)
y <- rep(0, times=4)
y[1] <- conf_mx$table[2]
y[2] <- conf_mx$table[1]
y[3] <- conf_mx$table[4]
y[4] <- conf_mx$table[3]
df <- data.frame(TClass, PClass, y)

p1 <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
  geom_tile(aes(fill = y), colour = "white") +
  geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_bw() + theme(legend.position = "none")

outfig <- file.path(figfolder, "reduced_data_best_conf_mx_normal_vs_stageI.pdf")
ggsave(outfig, plot=p1, width=4, height=4)
```

stage I from stage IV
```{r lower sample size stage I vs stage IV}
#gene_list <- c("RASIP1", "RASSF1", "ARHGAP20")
gene_list <- names(log_best_list[[2]]$coefficients)[
                       2:length(names(log_best_list[[2]]$coefficients))]
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(gene_list, "category_factor")]
# now select the given amount of samples
# option  1) 8 normal, 5 stage I, 10 stage IV
# option 2) 5 normal, 10 stage I, 6 stage IV
# randomly select these number of samples
normal_index <- which(ras_tmp$category_factor=="Normal")
normal_selected <- sample(normal_index, size=5)
stageI_index <- which(ras_tmp$category_factor=="Tumor_Stage_I")
stageI_selected <- sample(stageI_index, size=10)
stageIV_index <- which(ras_tmp$category_factor=="Tumor_Stage_IV")
stageIV_selected <- sample(stageIV_index, size=6)

ras_tmp2 <- ras_tmp[c(normal_selected, stageI_selected, stageIV_selected),]
# scale the new dataframe
ras_tmp2[, gene_list] <- scale(ras_tmp2[, gene_list])
# repeat the anova plots
plotlist <- list()
for (gene in gene_list){
  stagelm <- lm(as.formula(paste0(gene, " ~ category_factor")), data=ras_tmp2)
  stageav <- aov(stagelm)
  summary(stageav)
  tukeydf <- as_tibble(TukeyHSD(stageav, conf.level=.95)$category_factor, 
                       rownames="comparison")
  tukeydf$p.signif <- stars.pval(tukeydf$`p adj`)
  tukeydf$p.signif[tukeydf$p.signif==" "] <- "ns"
  tukeydf$group1 <- limma::strsplit2(tukeydf$comparison, split="-")[,2]
  tukeydf$group2 <- limma::strsplit2(tukeydf$comparison, split="-")[,1]
  
  # use Tukey later, but test with the example given
  tmpdf <- compare_means(as.formula(paste0(gene, " ~ category_factor")), data=ras_tmp)
  my_comparisons <- list( c("Normal", "Tumor_Stage_I"), 
                          c("Normal", "Tumor_Stage_IV"), 
                          c("Tumor_Stage_I", "Tumor_Stage_IV") )
  # set y positions on top of highest notch
  p1 <- ggplot(data=ras_tmp2)+
    geom_boxplot(mapping=aes(x=category_factor, y=.data[[gene]]))+
    scale_x_discrete(labels=c("Normal","Stage I","Stage IV"))+
    theme(axis.title.x=element_blank(),
          axis.text.x = element_text(angle = 45, vjust = 0.4, hjust=1),
          plot.margin = unit(c(0.05, 0.05, 0.3, 0.05), 
                                "inches"))
  ymax <- median(ras_tmp2[[gene]]) + 2 * IQR(ras_tmp2[[gene]])
  ymin <- median(ras_tmp2[[gene]]) - 2 * IQR(ras_tmp2[[gene]])
  ystep <- (ymax-ymin)/10
  p2 <- p1+ stat_pvalue_manual(tukeydf, label = "p.signif", 
    y.position = c(ymax+0.1*ystep, ymax+1.1*ystep, ymax+2.1*ystep))
  plotlist[[gene]] <- p2
}
# create a plot with normal, tumor and stage I
plot_all <- cowplot::plot_grid(plotlist=plotlist, nrow=1)
figfile <- file.path(figfolder, "best_genes_reduced_tukey_stageI_vs_stageIV.pdf")
ggsave(figfile, plot=plot_all, width=14, height=5)

# can we seperate tumor stage I from stage IV using these genes?
ras_tmp3 <- ras_tmp2[ras_tmp2$category_factor %in% c("Tumor_Stage_I", "Tumor_Stage_IV"),]
ras_tmp3$category_factor <- droplevels(ras_tmp3$category_factor)
it_model  <-  glm(as.formula(paste0("category_factor ~ ",
                                    paste(gene_list,collapse="+"))),
                  data = ras_data, family = "binomial")
conf_mx <- runCV(ras_tmp3, it_model,
                 depvar="category_factor",
                 depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
                 posclass="Tumor_Stage_IV",
                 return_conf_mx=T)
#print(conf_mx)
# plot conf_mx
TClass <- factor(c(0, 0, 1, 1))
PClass <- factor(c(0, 1, 0, 1))
# y <- c(FN, TP, TN, FP)
y <- rep(0, times=4)
y[1] <- conf_mx$table[2]
y[2] <- conf_mx$table[1]
y[3] <- conf_mx$table[4]
y[4] <- conf_mx$table[3]
df <- data.frame(TClass, PClass, y)

p2 <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
  geom_tile(aes(fill = y), colour = "white") +
  geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_bw() + theme(legend.position = "none")

outfig <- file.path(figfolder, "reduced_data_best_conf_mx_stageI_vs_stageIV.pdf")
ggsave(outfig, plot=p2, width=4, height=4)
```

Use bootstrapping to calculate the mean and sd of balanced accuracy
```{r lower sample size repeated with more genes normal vs stageI}
#gene_list <- c("PLCE1", "RGL1", "RAF1")
#gene_list <- c("PLCE1", "RGL2", "RAF1")
gene_list <- c("PLCE1", "RAF1")
test_char <- paste0(gene_list, collapse="_")
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(gene_list, "category_factor")]
# now select the given amount of samples
# option  1) 8 normal, 5 stage I, 10 stage IV
# option 2) 5 normal, 10 stage I, 6 stage IV
# randomly select these number of samples
num_boot <- 1000
acc_list <- list()
for (ii in 1:num_boot){
  if (ii %% 10==0) print(ii)
  normal_index <- which(ras_tmp$category_factor=="Normal")
  normal_selected <- sample(normal_index, size=5, replace=T)
  stageI_index <- which(ras_tmp$category_factor=="Tumor_Stage_I")
  stageI_selected <- sample(stageI_index, size=10, replace=T)
  stageIV_index <- which(ras_tmp$category_factor=="Tumor_Stage_IV")
  stageIV_selected <- sample(stageIV_index, size=6, replace=T)
  
  ras_tmp2 <- ras_tmp[c(normal_selected, stageI_selected, stageIV_selected),]
  # scale the new dataframe
  ras_tmp2[, gene_list] <- scale(ras_tmp2[, gene_list])
  # can we seperate normal from tumor stage I using these genes?
  ras_tmp3 <- ras_tmp2[ras_tmp2$category_factor %in% c("Normal", "Tumor_Stage_I"),]
  ras_tmp3$category_factor <- droplevels(ras_tmp3$category_factor)
  it_model  <-  glm(as.formula(paste0("category_factor ~ ",
                                      paste(gene_list,collapse="+"))),
                    data = ras_tmp3, family = "binomial")
  acc <- runCV_LOOCV(ras_tmp3, it_model,
                   depvar="category_factor",
                   depvar_levels=c("Normal", "Tumor_Stage_I"),
                   posclass="Tumor_Stage_I",
                   return_conf_mx=F)
  acc_list[[ii]] <- acc
}
ba_vector_normal_vs_stageI <- unlist(acc_list)
ptitle <- paste0(it_model$formula)[3]
p <- ggplot(data=as.data.frame(ba_vector_normal_vs_stageI), 
            mapping=aes(x=ba_vector_normal_vs_stageI))+
  geom_histogram(binwidth=0.05)+
  xlab("balanced accuracy")+
  ggtitle(ptitle)
suppressWarnings(plot(p))
figfile <- file.path(figfolder, 
        paste("histogram_bootsrap_normal_vs_stageI_", test_char,".pdf"))
suppressWarnings(ggsave(figfile, p, width=3, height=3))
```

```{r lower sample size repeated with more genes stageI vs stageIV}
# make sure the gene list is obtained from running BIC for stage I vs stage IV
# still TO DO properly!!!
gene_list <- names(
log_best_list_stageI_vs_stageIV <- log_best_list[[2]]$coefficients)[
                       2:length(names(log_best_list[[2]]$coefficients))]
#gene_list <- c("RGL2", "RASSF1", "RADIL", "RASSF7", "RAPGEF4", "KRAS", "RIN3")
# gene_list <- c("RGL2", "RASSF1", "RASSF7", "KRAS")
# gene_list <- c("NRAS", "PIK3CA", "RALGDS", "RGL2", "RASSF1", "RASSF7", 
#                   "MLLT4")
test_char <- paste0(gene_list, collapse="_")
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Normal", "Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(gene_list, "category_factor")]
# now select the given amount of samples
# option  1) 8 normal, 5 stage I, 10 stage IV (104)
# option 3) 5 normal, 10 stage I, 6 stage IV (102)
# randomly select these number of samples
num_boot <- 10000
acc_list <- list()
for (ii in 1:num_boot){
  if (ii %% 100==0) print(ii)
  normal_index <- which(ras_tmp$category_factor=="Normal")
  normal_selected <- sample(normal_index, size=5, replace=T)
  stageI_index <- which(ras_tmp$category_factor=="Tumor_Stage_I")
  stageI_selected <- sample(stageI_index, size=10, replace=T)
  stageIV_index <- which(ras_tmp$category_factor=="Tumor_Stage_IV")
  stageIV_selected <- sample(stageIV_index, size=6, replace=T)
  
  ras_tmp2 <- ras_tmp[c(normal_selected, stageI_selected, stageIV_selected),]
  # scale the new dataframe
  ras_tmp2[, gene_list] <- scale(ras_tmp2[, gene_list])
  # can we seperate tumor stage I from stage IV using these genes?
  ras_tmp3 <- ras_tmp2[ras_tmp2$category_factor %in% c("Tumor_Stage_I", "Tumor_Stage_IV"),]
  ras_tmp3$category_factor <- droplevels(ras_tmp3$category_factor)
  it_model  <-  glm(as.formula(paste0("category_factor ~ ",
                                      paste(gene_list,collapse="+"))),
                    data = ras_tmp3, family = "binomial")
  acc <- runCV_LOOCV(ras_tmp3, it_model,
                   depvar="category_factor",
                   depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
                   posclass="Tumor_Stage_IV",
                   return_conf_mx=F)
  acc_list[[ii]] <- acc
}
ba_vector_stageI_vs_stageIV <- unlist(acc_list)
ptitle <- paste0(it_model$formula)[3]
pmedian <- median(ba_vector_stageI_vs_stageIV)
p <- ggplot(data=as.data.frame(ba_vector_stageI_vs_stageIV), 
            mapping=aes(x=ba_vector_stageI_vs_stageIV))+
  geom_histogram(binwidth=0.05)+
  geom_vline(xintercept=pmedian, color="red")+
  xlab("balanced accuracy")+
  ggtitle(ptitle)
suppressWarnings(plot(p))
figfile <- file.path(figfolder, 
        paste("histogram_bootsrap_stageI_vs_stageIV_", test_char,".pdf"))
suppressWarnings(ggsave(figfile, p, width=6, height=3))

# plot results together?
```

# Finally reduce the two-gene accuracy/aic/bic excel tables to contain only the reduced genes
```{r reduce existing tables}
filename1 <- "best_model_2genes_normal_vs_stageI.xlsx"
filename2 <- "best_model_2genes_stageI_vs_stageIV.xlsx"
for (filename in c(filename1, filename2)){
  newfile <- file.path(phome, "Results" ,filename)
  if (!file.exists(newfile)){
    oldfile <-  file.path(phome, "Results", "archive_new", filename)
    oldtable <- read_excel(oldfile, sheet=rtype)
    omitrows <- c()
    for (ri in 1:nrow(oldtable)){
      if (oldtable$gene1[ri] %in% exclude | oldtable$gene2[ri] %in% exclude){
        omitrows <- c(omitrows, ri)
      }
    }
    newtable <- oldtable[-omitrows,]
    outlist <- list()
    outlist[[rtype]] <- newtable
    openxlsx::write.xlsx(outlist, file=newfile)
  }
}
```

# It was calculated seperately on the server, but based on the best 
obtained accuracy values, show the best model seperating
tumor stageI from stageIV and write out corresponding metrics
```{r best optimized model}
ras_tmp <- ras_data[ras_data$category2 %in% 
                      c("Tumor_Stage_I", "Tumor_Stage_IV"),
                    c(indep_vars, "category_factor")]
# also delete unused factors
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)
# these genes were obtained by the optimization
# test_genes <- c("RGL2", "RASSF1", "RADIL", "RASSF7", "RAPGEF4", "KRAS", "RIN3",
#                 "PLCE1", "RAF1")
test_genes <- c("RGL2", "RASSF1", "RADIL", "RASSF7", "RAPGEF4", "KRAS", "RIN3")
test_char <- paste0(test_genes, collapse="_")
it_model  <-  glm(as.formula(paste0("category_factor ~ ",
                                    paste(test_genes,collapse="+"))),
                  data = ras_tmp, family = "binomial")
red_conf_mx <- runCV_LOOCV(ras_tmp, it_model,
         depvar="category_factor",
         depvar_levels=c("Tumor_Stage_I", "Tumor_Stage_IV"),
         posclass="Tumor_Stage_IV",
         return_conf_mx=T)
print(red_conf_mx)
# also plot
  conf_mx <- red_conf_mx
  TClass <- factor(c(0, 0, 1, 1))
  PClass <- factor(c(0, 1, 0, 1))
  # y <- c(FN, TP, TN, FP)
  y <- rep(0, times=4)
  y[1] <- conf_mx$table[2]
  y[2] <- conf_mx$table[1]
  y[3] <- conf_mx$table[4]
  y[4] <- conf_mx$table[3]
  df <- data.frame(TClass, PClass, y)
  
  p <- ggplot(data =  df, mapping = aes(x = TClass, y = PClass)) +
    geom_tile(aes(fill = y), colour = "white") +
    geom_text(aes(label = sprintf("%1.0f", y)), vjust = 1, size=10) +
    scale_fill_gradient(low = "blue", high = "red") +
    theme_bw() + theme(legend.position = "none")
  
  plot(p)
  print(conf_mx)
  outfig <- file.path(figfolder, 
                      paste("best_conf_mx_", test_char, ".pdf", sep=""))
  ggsave(outfig, plot=p, width=7, height=7)


# also seperate tumor stage I from stage IV
indata_mx <-  as.matrix(ras_tmp[, test_genes])
lda_model <- MASS::lda(indata_mx, grouping=ras_tmp$category_factor)
lda_values <- predict(lda_model, newdata = ras_tmp[, test_genes])
all.equal(lda_values$class, ras_tmp$category_factor) # "14 string mismatches"
lda_annot <- as_tibble(lda_values$x)
lda_annot$category_factor <- ras_tmp$category_factor

# obtain the “proportion of trace” which is the proportion of between-class variance that is explained by successive discriminant functions. 
prop_trace <- prop.table(lda_model$svd^2)
nicelabel <- list()
for (ii in 1:length(prop_trace)){
 nicelabel[[ii]] <- paste('LD',
        as.character(ii), ' (', 
        as.character(round(100*prop_trace[ii], digits = 2)), '%)', sep="")
}
p <- ggplot(data=lda_annot,  mapping=aes(x=LD1, fill=category_factor)) +
  geom_histogram()

# plot density curve for all stages
lda_values <- predict(lda_model, newdata = ras_data[, test_genes])
lda_annot <- as_tibble(lda_values$x)
lda_annot$sample <- ras_data$sample
lda_annot$category_factor <- ras_data$category_factor

lda_tmp <- lda_annot[!lda_annot$category_factor %in% 
                       c("Normal", "Tumor_Stage_X"),]
lda_tmp$category_factor <- droplevels(lda_tmp$category_factor)

p <- ggplot(data=lda_tmp, mapping=aes(x=LD1, color=category_factor))+
  geom_density(fill="white",  alpha=0.25)+
      scale_color_manual("category",
            values = c("Normal" = "black",
                       "Tumor_Stage_I"="navy",
                       "Tumor_Stage_II"="purple4",
                       "Tumor_Stage_III"="purple",
                       "Tumor_Stage_IV"="red"),
            labels= c("Normal" = "Normal",
                       "Tumor_Stage_I"="Tumor Stage I",
                       "Tumor_Stage_II"="Tumor Stage II",
                       "Tumor_Stage_III"="Tumor Stage III",
                       "Tumor_Stage_IV"="Tumor Stage IV"))
plot(p)
outfig <- file.path(figfolder, 
                    paste("ras_all_lda_density",test_char,".pdf", sep=""))
ggsave(outfig, plot=p, width=7, height=7)
```

