# 2024.10.01.
# Anna Lovrics

---
title: "Ras interactome analysis"
date: "`r Sys.Date()`"
author: "Anna Lovrics"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```


```{r libraries}
library(readxl) # another package to read xls files
library(dplyr)
library(tidyr) # to transform data for ggplot2 plots
library(data.table) # while I use tibble mostly, fread is best input option
library(tibble)
library(ggplot2) # for visualization
library(EnvStats) # for statistic tests
library(ggcorrplot) # for nice correlation plots
library(cowplot) # to plot q-q plots in a grid
library(gtools) # specifically to generate stars from significance values
library(caret) # for Classification And REgression Training
library(pROC) # to easily plot ROC curves
library(limma) # for the strsplit2 function
library(gridExtra) # to plot a list of figures in a grid
library(ggpubr) # for the same purpose to arrange figures in a grid and also for significance stars
library(arsenal) # Functions for Large-Scale Statistical Summaries
```

Yes, the method you described can be a reasonable approach to assess the separability of two sample classes, provided that the principal components are calculated appropriately and the criterion for "separation" is well-defined. Here’s an explanation of how this approach works and some considerations:

---

### **Rationale**
1. **Principal Component Analysis (PCA)**:
   - PCA reduces the dimensionality of data by finding orthogonal axes (principal components) that maximize variance.
   - The first few principal components capture the majority of the variance in the data.

2. **Fewer Components for Separation**:
   - If two sample classes can be separated in fewer principal components, this indicates that their variance is largely aligned with the main axes of variation. This typically means they are more linearly separable in the reduced-dimensional space.

3. **Comparison Criterion**:
   - By counting how many principal components are required to achieve a certain level of separation, you create a metric for separability: **fewer components = greater separability**.

---

### **Steps to Implement**
1. **Perform PCA on the Dataset**:
   - Compute the principal components for the combined dataset of both classes.

2. **Project Data onto Principal Components**:
   - Represent the data using the first `k` principal components (e.g., starting with k=1).

3. **Define Separation Criterion**:
   - For example:
     - Visual Inspection: Plot the first `k` components and check if the classes are visually separable.
     - Quantitative Measures:
       - Compute classification accuracy using a simple classifier (e.g., linear SVM or logistic regression) in the reduced `k`-dimensional space.
       - Calculate metrics like the Silhouette Score or cluster separation indices.
       - Use pairwise distance metrics to quantify separation.

4. **Determine the Minimum Number of Components Needed**:
   - Identify the smallest number of principal components `k` such that the classes are considered “separated” based on your chosen criterion.

5. **Compare Across Classes**:
   - Repeat this process for different pairs of classes. Classes that require fewer components for separation are more separable.

---

### **Considerations and Limitations**
1. **Choice of Separation Criterion**:
   - Ensure that the criterion for "separation" is clear and consistent (e.g., specific threshold for classification accuracy, Silhouette Score, etc.).

2. **Linearity**:
   - PCA finds linear combinations of features. If the separation between classes is nonlinear, this method may underestimate separability.

3. **Dimensionality of the Dataset**:
   - If the dataset has very high dimensionality, PCA might require more components simply due to noise or irrelevant features, rather than true overlap between classes.

4. **Class Imbalance**:
   - If one class dominates, separation may appear easier or harder depending on the projection. Consider stratification or weighting during analysis.

5. **Robustness Testing**:
   - Validate the results using cross-validation or by splitting the data into training and testing sets to avoid overfitting.

---

### **Extensions**
- **Nonlinear Dimensionality Reduction**:
  - If the classes are not linearly separable, consider using nonlinear methods like Kernel PCA, t-SNE, or UMAP to explore separability in a different space.

- **Statistical Testing**:
  - Test whether the observed separability (e.g., number of components) is statistically significant using permutation tests.

- **Feature Selection**:
  - Examine which features contribute most to the separability in the principal components.

---

### **Summary**
Your proposed method is valid and interpretable for assessing separability when the separation is largely linear. By counting the number of principal components required for separation, you effectively quantify how well the main axes of variance in the data align with the distinctions between classes. However, it’s essential to choose an appropriate separation criterion and be aware of potential nonlinearities or other limitations in the data.

# Folder structure
```{r folders}
phome <- "/media/anna/bigdisk/Anna/Enzimhome_local/Projects/StatisticsCore/Ras"
datafolder <- file.path(phome, "Data")
figfolder <- file.path(phome, "Figures")
resfolder <- file.path(phome, "Results")
if (!file.exists(figfolder)) dir.create(figfolder)
calcfolder <- file.path(phome, "Data_calculated")
if (!file.exists(calcfolder)) dir.create(calcfolder)
# source content of utility file
rscriptfolder <- file.path(phome, "Rscript")
util_file <- file.path(rscriptfolder, "anal_utils.R")
source(util_file)
```

# define the cancer type
```{r cancer type}
#rtype <- "COAD"
rtype <- "COAD"
figfolder <- file.path(figfolder, rtype)
if (!file.exists(figfolder)) dir.create(figfolder)
calcfolder <- file.path(calcfolder, rtype)
if (!file.exists(calcfolder)) dir.create(calcfolder)
resfolder <- file.path(resfolder, rtype)
if (!file.exists(resfolder)) dir.create(resfolder)
```

# read in data
```{r read data, results="asis"}
ras_data <- fread(file.path(datafolder, 
                  paste("TCGA_", rtype, "_plus_stage_20241028.tsv", sep="")),
                     sep="\t")
ras_data <- as_tibble(ras_data)
# remove FFPE Scrolls sample
tmpi <- which(ras_data$sample_type=="FFPE Scrolls")
if (length(tmpi) > 0) ras_data <- ras_data[-tmpi,]
# exclude some of the genes
exclude <- c("RAPGEF3", "RAPGEF5", "RASIP1", "MYO9B", "PIK3C2A", "RGL3", "RGL4",
             "TIAM1", "ARHGAP20","RASSF2", "RASSF3", "RASSF4", "RASSF6")
keep_cols <- setdiff(colnames(ras_data), exclude)
ras_data <- ras_data[, keep_cols]
# View(ras_data)
# first two columns are the same -> remove
# all.equal(ras_data$sample, ras_data$samples)
ras_data$samples <- NULL
indep_vars <- setdiff(colnames(ras_data), 
                      c( "sample","pathologic_stage","sample_type"))
# keep original data, replace columns with what we need in the computations
ras_data_orig <- ras_data
# add a column for tumor or normal
ras_data$sample_type2 <- ras_data$sample_type
ras_data$sample_type2[ras_data$sample_type2 %in% 
    c("Metastatic", "Primary Tumor", "Recurrent Tumor")] <- "Tumor"
ras_data$sample_type2[ras_data$sample_type2=="Solid Tissue Normal"] <- "Normal"
# add a column for stages I-IV or X
ras_data$pathologic_stage2 <- ras_data$pathologic_stage
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage I","Stage IA", "Stage IB", "Stage IC")] <- "Stage_I"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage II","Stage IIA", "Stage IIB", "Stage IIC")] <- "Stage_II"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage III","Stage IIIA", "Stage IIIB", "Stage IIIC")] <- "Stage_III"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("Stage IV", "Stage IVA", "Stage IVB", "Tumor_Stage IVB")] <- "Stage_IV"
ras_data$pathologic_stage2[ras_data$pathologic_stage2 %in% 
    c("[Discrepancy]","Stage X")] <- "Stage_X"

# sample type plusz stage together
ras_data$category <- paste(ras_data$sample_type2, ras_data$pathologic_stage2, sep="_")
ras_data$category2 <- ras_data$category
ras_data$category2[ras_data$category2 %in% c("Normal_Stage_X", "Normal_Stage_I",
     "Normal_Stage_II", "Normal_Stage_III", "Normal_Stage_IV")] <- "Normal"

# check for constant columns and remove if any
cvar <- apply(ras_data[, indep_vars], 2, var, na.rm=T)
ci <- which(cvar==0)
if (length(ci)>0) ras_data <- ras_data[-ci,]

# add some more response variables
ras_data$sample_type_factor <- factor(ras_data$sample_type2,
                                 levels=c("Normal", "Tumor"))
ras_data$sample_type_bool <- ifelse(ras_data$sample_type2=="Tumor", F, T)
ras_data$category_factor <- factor(ras_data$category2)
# save data
saveRDS(list("data"=ras_data, "indep_vars"=indep_vars),
        file=file.path(calcfolder, paste("TCGA_", rtype, ".rds", sep="")))

# # print out tables for check
# print(table(ras_data$sample_type_factor))
# print(table(ras_data$category_factor))
# xtable::xtable(table(ras_data$category_factor))
```

# test how many PCA components are needed to seperate data
test for separation: k-NN
```{r PCA}
balanced_accuracy_tbl1 <- list()
balanced_accuracy_tbl2 <- list()
depvar <- "category_factor"
depvar_levels <- c("Normal", "Tumor_Stage_I")
depvar_levels_str <- paste0(depvar_levels, collapse='_vs_')
ri <- which(ras_data$category2 %in% depvar_levels)
ras_tmp <- ras_data[ri, ]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)

indata_mx <- as.matrix(ras_tmp[, indep_vars])
rownames(indata_mx) <- ras_tmp$sample
pca <- prcomp(indata_mx, center=T, scale=T,
              tol = sqrt(.Machine$double.eps), rank.=30)
# scree plot
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

sumvar <- 0
maxi <- 1
for (pcind in 1:length(pca.var)){
  sumvar <- sumvar + pca.var.per[pcind]
  #print(c(pcind, sumvar))
  if (sumvar > 95){
    # if (sumvar >= 100){
    maxi <- pcind
    #print(maxi)
    break
  }
}
vardf <- as.data.frame(cbind('PC'=1:maxi, 'variance'=pca.var.per[1:maxi]))

p <- ggplot(data=vardf) +
  geom_col(mapping=aes(x=PC, y=cumsum(variance)))+
  xlab("Principal Component") +  ylab("Cumulative percent variation") +
  geom_hline(yintercept=95, linetype="dashed",
             color = "red", linewidth=1)
plot(p)
# save scree plot as well
figfile <- file.path(figfolder, 
            paste("ras_all_scree_plot_", depvar_levels_str,".pdf", sep=""))
ggsave(figfile, plot=p, height=5, width=5)
# now plot a PCA plot with annotation
pca_res <- as_tibble(pca$x, rownames = "sample")
pca_annot <- inner_join(pca_res, ras_data, by=c("sample"))
pca_annot <- droplevels(pca_annot)

for (kval in 1:10){
 balanced_acccuracy_vector <- list()
for (ii in 1:nrow(vardf)){
  pca_tmp <- inner_join(pca_res[,1:(ii+1)], 
                        ras_data[, c(depvar, "sample")],
                        by = join_by(sample))
  pca_tmp[[depvar]] <- droplevels(pca_tmp[[depvar]])
  balanced_acccuracy_vector[[ii]] <- runCV_LOOCV_kNN(pca_tmp, 
                  depvar=depvar,
                  depvar_levels=depvar_levels,
                  posclass=depvar_levels[[2]],
                  kval=kval,
                  verbose=F,
                 return_conf_mx=F)
}
balanced_acccuracy_vector <- unlist(balanced_acccuracy_vector)
balanced_accuracy_tbl <- as_tibble(as.data.frame(balanced_acccuracy_vector),
                                   rownames="index")
balanced_accuracy_tbl$index <- as.numeric(balanced_accuracy_tbl$index)
balanced_accuracy_tbl1[[kval]] <- balanced_accuracy_tbl
p <- ggplot(data=balanced_accuracy_tbl,
            mapping=aes(x=index, y=balanced_acccuracy_vector))+
     geom_point()+ geom_line()
outfile <- file.path(figfolder, 
            paste("kNN_pca_plot_k_", as.character(kval), 
            depvar_levels_str,".pdf", sep=""))
ggsave(outfile, plot=p, width=7, height=7)  
}

# keep the first 6 PCs only
vardf <- vardf[1:6,]
nicelabel <- list()
for (ii in 1:nrow(vardf)){
 nicelabel[[ii]] <- paste('PC',
        as.character(ii), ' (', as.character(vardf$variance[ii]), '%)', sep="")
}

plist <- list()
for (ii in 1:(nrow(vardf)/2)){
  xname <- colnames(pca_annot)[2*ii]
  yname <- colnames(pca_annot)[2*ii+1]
  plist[[ii]] <- ggplot(data=pca_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           col  = category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category", 
                       values = c("Normal" = "black", "Tumor_Stage_I" = "navy"))+
    labs(x = nicelabel[[2*(ii-1)+1]], y = nicelabel[[2*ii]])+
    #geom_text(mapping=aes(x = PC1, y = PC2, label=samples), size=1)+
    theme_gray(base_size = 20) # change the default font size in the grey theme  
}

pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 2)
resfile <- file.path(figfolder, 
       paste("ras_all_pca_plot_", depvar_levels_str,".pdf", sep=""))
ggsave(resfile, plot=pfinal, width=20, height=12)


################################
depvar <- "category_factor"
depvar_levels <- c("Tumor_Stage_I", "Tumor_Stage_IV")
depvar_levels_str <- paste0(depvar_levels, collapse='_vs_')
ri <- which(ras_data$category2 %in% depvar_levels)
ras_tmp <- ras_data[ri, ]
ras_tmp$category_factor <- droplevels(ras_tmp$category_factor)

indata_mx <- as.matrix(ras_tmp[, indep_vars])
rownames(indata_mx) <- ras_tmp$sample
pca <- prcomp(indata_mx, center=T, scale=T,
              tol = sqrt(.Machine$double.eps), rank.=30)
# scree plot
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

sumvar <- 0
maxi <- 1
for (pcind in 1:length(pca.var)){
  sumvar <- sumvar + pca.var.per[pcind]
  #print(c(pcind, sumvar))
  if (sumvar > 95){
    # if (sumvar >= 100){
    maxi <- pcind
    #print(maxi)
    break
  }
}
vardf <- as.data.frame(cbind('PC'=1:maxi, 'variance'=pca.var.per[1:maxi]))

p <- ggplot(data=vardf) +
  geom_col(mapping=aes(x=PC, y=cumsum(variance)))+
  xlab("Principal Component") +  ylab("Cumulative percent variation") +
  geom_hline(yintercept=95, linetype="dashed",
             color = "red", linewidth=1)
plot(p)
# save scree plot as well
figfile <- file.path(figfolder, 
            paste("ras_all_scree_plot_", depvar_levels_str,".pdf", sep=""))
ggsave(figfile, plot=p, height=5, width=5)
# now plot a PCA plot with annotation
pca_res <- as_tibble(pca$x, rownames = "sample")
pca_annot <- inner_join(pca_res, ras_data, by=c("sample"))
pca_annot <- droplevels(pca_annot)

for (kval in 1:10){
 balanced_acccuracy_vector <- list()
for (ii in 1:nrow(vardf)){
  pca_tmp <- inner_join(pca_res[,1:(ii+1)], 
                        ras_data[, c(depvar, "sample")],
                        by = join_by(sample))
  pca_tmp[[depvar]] <- droplevels(pca_tmp[[depvar]])
  balanced_acccuracy_vector[[ii]] <- runCV_LOOCV_kNN(pca_tmp, 
                  depvar=depvar,
                  depvar_levels=depvar_levels,
                  posclass=depvar_levels[[2]],
                  kval=kval,
                  verbose=F,
                 return_conf_mx=F)
}
balanced_acccuracy_vector <- unlist(balanced_acccuracy_vector)
balanced_accuracy_tbl <- as_tibble(as.data.frame(balanced_acccuracy_vector),
                                   rownames="index")
balanced_accuracy_tbl$index <- as.numeric(balanced_accuracy_tbl$index)
balanced_accuracy_tbl2[[kval]] <- balanced_accuracy_tbl
p <- ggplot(data=balanced_accuracy_tbl,
            mapping=aes(x=index, y=balanced_acccuracy_vector))+
     geom_point()+ geom_line()
outfile <- file.path(figfolder, 
            paste("kNN_pca_plot_k_", as.character(kval), 
            depvar_levels_str,".pdf", sep=""))
ggsave(outfile, plot=p, width=7, height=7)  
}

# keep the first 6 PCs only
vardf <- vardf[1:6,]
nicelabel <- list()
for (ii in 1:nrow(vardf)){
 nicelabel[[ii]] <- paste('PC',
        as.character(ii), ' (', as.character(vardf$variance[ii]), '%)', sep="")
}

plist <- list()
for (ii in 1:(nrow(vardf)/2)){
  xname <- colnames(pca_annot)[2*ii]
  yname <- colnames(pca_annot)[2*ii+1]
  plist[[ii]] <- ggplot(data=pca_annot) +
    geom_point(mapping=aes(x = .data[[xname]], y = .data[[yname]], 
                           col  = category_factor),
               size = 1, stroke = 1, fill="white")+
    # color points as required
    scale_color_manual("category", 
                values = c("Tumor_Stage_I" = "navy", "Tumor_Stage_IV" = "red"))+
    labs(x = nicelabel[[2*(ii-1)+1]], y = nicelabel[[2*ii]])+
    #geom_text(mapping=aes(x = PC1, y = PC2, label=samples), size=1)+
    theme_gray(base_size = 20) # change the default font size in the grey theme  
}

pfinal <- cowplot::plot_grid(plotlist=plist, ncol = 2)
resfile <- file.path(figfolder, 
              paste("ras_all_pca_plot_", depvar_levels_str,".pdf", sep=""))
ggsave(resfile, plot=pfinal, width=20, height=12)

for (kval in 1:length(balanced_accuracy_tbl1)){
  tbl1 <- balanced_accuracy_tbl1[[kval]]
  colnames(tbl1) <- c("index", "Normal_vs_Tumor_Stage_I")
  tbl2 <- balanced_accuracy_tbl2[[kval]]
  colnames(tbl2) <- c("index", "Tumor_Stage_I_vs_Tumor_Stage_IV")
  tbl_all <-  inner_join(tbl1, tbl2, by = join_by(index))
  p <- ggplot(data=tbl_all)+
     geom_point(mapping=aes(x=index, y=Normal_vs_Tumor_Stage_I))+ 
     geom_line(mapping=aes(x=index, y=Normal_vs_Tumor_Stage_I))+
     geom_point(mapping=aes(x=index, y=Tumor_Stage_I_vs_Tumor_Stage_IV),
                color="blue")+ 
     geom_line(mapping=aes(x=index, y=Tumor_Stage_I_vs_Tumor_Stage_IV),
                 color="blue")+
     geom_hline(yintercept = 0.5, color="red")+
     ylab("balanced accuracy") + xlab("PC index")
  outfile <- file.path(figfolder, 
              paste("kNN_pca_plot_k_", as.character(kval), 
              "_both.pdf", sep=""))
  ggsave(outfile, plot=p, width=7, height=7)  
}
```



